---
title: "Practice 9 Answers"
subtitle: "PUBH 523/623"
author: "Nicky Wakim"
date-modified: "today"
format: 
  html:
    link-external-newwindow: true
    toc: true
    self-contained: true
    page-layout: full
    html-math-method: mathjax
    self-contained-math: true
    highlight-style: ayu
    code-block-bg: "#f8f9fa"   
    code-block-border-left: "#5BAFF8" 
editor_options: 
  chunk_output_type: console
---

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#| message: false
#| warning: false
#| echo: false

pacman::p_load(
  tidyverse, 
  here, 
  rio,
  naniar,
  janitor,
  gt
)
```

## Questions

### Question 1: Missingness statistics

In Lesson 34, we saw how to get quick, high-level counts of missingness before doing anything else with a dataset.

#### Part A: Import dataset

::: {.callout-warning appearance="minimal"}
##### Task

Import `hrs_data.rds` and assign it to `hrs_data`.
:::

::: blue
**Answer:**

Here's an example to show you the pattern — the rest of your answers should follow this same style, with code chunks and a `#| ...` chunk options as needed.

```{r}
hrs_data <- import(here("data", "hrs_data.rds"))

glimpse(hrs_data)
```
:::

#### Part B: Total missingness in the dataset

::: {.callout-warning appearance="minimal"}
##### Task

Use `n_miss()` on `hrs_data` to find the total number of missing cells across the entire dataset.
:::

::: blue
**Answer:**

Not given
:::

#### Part C: Proportion of missingness in the dataset

::: {.callout-warning appearance="minimal"}
##### Task

Use `prop_miss()` on `hrs_data` to find the proportion of all cells in the dataset that are missing.
:::

::: blue
**Answer:**

```{r}
prop_miss(hrs_data)
```
:::

#### Part D: Missingness in a single variable

::: {.callout-warning appearance="minimal"}
##### Task

Use `n_miss()` on just the `srh` column of `hrs_data` to find how many values are missing for self-reported health.
:::

::: {.callout-tip appearance="minimal"}
##### Tip

You can pull out a single column with `$`, e.g. `hrs_data$srh`.
:::

::: blue
**Answer:**

Not given
:::

#### Part E: Missingness summary table

::: {.callout-warning appearance="minimal"}
##### Task

Use `miss_var_summary()` on `hrs_data` to get one row per variable with its count and percent missing. Pipe the result into `gt()` to make it a nicer table. Assign the table to `hrs_01_summary_table` and display it.

Which variable has the most missing values? Which has the least?
:::

::: blue
**Answer:**

Not given
:::

### Question 2: Visualizing missingness

In Lesson 34, we also covered a few plotting functions from `naniar` that help us see *where* and *how* data is missing.

#### Part A: A bird's-eye view with `vis_miss()`

::: {.callout-warning appearance="minimal"}
##### Task

Use `vis_miss()` on `hrs_data` to visualize missingness across the whole dataset. Assign the plot to `hrs_02_plot_vis_miss` and display it.
:::

::: blue
**Answer:**

Start with something like:

```{r}
#| eval: false

hrs_02_plot_vis_miss <- hrs_data |> 
  ______()
```
:::

#### Part B: Customize the plot

::: {.callout-warning appearance="minimal"}
##### Task

`vis_miss()` returns a `ggplot2` object, so you can add layers to it just like the plots you built last week. Take `hrs_02_plot_vis_miss` and add a `theme()` layer that rotates the x-axis text labels by 20 degrees (`angle = 20`) so they're easier to read. Assign the result to `hrs_03_plot_vis_miss_clean` and display it.

Looking at your plot, is missingness concentrated in a few variables, spread across many observations, or both?
:::

::: {.callout-tip appearance="minimal"}
##### Tip

`theme(axis.text.x = element_text(angle = 20, hjust = 0))` adjusts the angle and alignment of the x-axis text.
:::

::: blue
**Answer:**

Not given
:::

#### Part C: Missingness by group with `gg_miss_fct()`

::: {.callout-warning appearance="minimal"}
##### Task

Use `gg_miss_fct()` on `hrs_data`, stratifying by `degree` (set `fct = degree`). Assign the plot to `hrs_04_plot_fct` and display it.

Does missingness look similar across `degree`, or does one group have noticeably more missing data for certain variables?
:::

::: blue
**Answer:**

Not given
:::

#### Part D: Missingness intersections with `gg_miss_upset()`

::: {.callout-warning appearance="minimal"}
##### Task

Use `gg_miss_upset()` on `hrs_data`, setting `nsets = 10` and `text.scale = 2`. Assign the plot to `hrs_05_plot_upset` and display it.

Which combination of variables is missing together most often? Is there a variable that tends to be missing mostly on its own?
:::

::: blue
**Answer:**

Not given
:::

### Question 3: Filtering out missing values

In Lesson 35, we covered how to remove rows with missing data, and a common gotcha when using `filter()` on a column with `NA`s.

#### Part A: Check missingness before you drop anything

::: {.callout-warning appearance="minimal"}
##### Task

Before dropping any rows, pipe `hrs_data` into `miss_var_summary()` and then `gt()` again, just like Question 1 Part E, so you have a "before" table to compare to later. Assign it to `hrs_06_summary_before`.
:::

::: {.callout-tip appearance="minimal"}
##### Tip

Getting in the habit of checking missingness *before* and *after* you drop or fill in values helps you catch mistakes, like accidentally dropping more rows than you meant to.
:::

::: blue
**Answer:**

Not given
:::

#### Part B: Drop rows missing anything

::: {.callout-warning appearance="minimal"}
##### Task

Use `drop_na()` on `hrs_data` (with no arguments) to remove any row that has *any* missing value. Assign the result to `hrs_07_dropna_all`. How many rows are left compared to `hrs_data`?
:::

::: blue
**Answer:**

Not given
:::

#### Part C: Drop rows missing specific columns

::: {.callout-warning appearance="minimal"}
##### Task

Now use `drop_na()` on `hrs_data` again, but this time only target the `act_vig` and `bp` columns, so rows are only dropped if one of *those two* columns is missing. Assign the result to `hrs_08_dropna_cols`. How does the number of rows compare to `hrs_07_dropna_all`?
:::

::: blue
**Answer:**

Start with something like:

```{r}
#| eval: false

hrs_08_dropna_cols <- hrs_data |> 
  drop_na(______, ______)
```
:::

#### Part D: The `filter()` gotcha

::: {.callout-warning appearance="minimal"}
##### Task

Use `tabyl()` (from `janitor`) on `hrs_data$smoke_ever` to see how many missing values `smoke_ever` has and confirm its categories.

Then, use `filter()` on `hrs_data` to keep only rows where `smoke_ever != "No"`. Assign the result to `hrs_09_filter_gotcha`. Use `tabyl()` on the `smoke_ever` column of `hrs_09_filter_gotcha` to check what happened to the missing values.
:::

::: blue
**Answer:**

Not given
:::

#### Part E: Fixing the gotcha with `filter_out()`

::: {.callout-warning appearance="minimal"}
##### Task

Redo Part D, but use `filter_out()` instead of `filter()` to exclude `smoke_ever == "No"` while keeping the `NA`s. Assign the result to `hrs_10_filter_out_fixed`. Confirm with `tabyl()` that the missing values are still there this time.
:::

::: blue
**Answer:**

Not given
:::

### Question 4: Filling in and recoding missing values

In Lesson 35, we also covered how to fill in missing values instead of dropping them, and how to turn a specific value into `NA` when needed.

#### Part A: Fill in a numeric variable

::: {.callout-warning appearance="minimal"}
##### Task

Use `mutate()` and `replace_na()` to create a new column, `cesd_filled`, in `hrs_data` that replaces any missing `cesd` (depression symptom score) with the mean `cesd` (use `na.rm = TRUE` inside `mean()`). Assign the result to `hrs_11_cesd_filled`.

Is filling in a missing value with the overall mean always a good idea? Why or why not?
:::

::: blue
**Answer:**

Start with something like:

```{r}
#| eval: false

hrs_11_cesd_filled <- hrs_data |> 
  mutate(
    cesd_filled = replace_na(______, ______)
  )
```
:::

#### Part B: Fill in a factor variable

::: {.callout-warning appearance="minimal"}
##### Task

Starting from `hrs_11_cesd_filled`, use `mutate()` and `fct_na_value_to_level()` to create a new column, `degree_filled`, that replaces any missing `degree` with a new level called `"Unknown"`. Assign the result to `hrs_12_degree_filled`.
:::

::: {.callout-tip appearance="minimal"}
##### Tip

`replace_na()` does not work on factor columns. Use `fct_na_value_to_level()` instead whenever the column you're filling in is a factor.
:::

::: blue
**Answer:**

Start with something like:

```{r}
#| eval: false

hrs_12_degree_filled <- hrs_11_cesd_filled |> 
  mutate(
    degree_filled = fct_na_value_to_level(______, level = ______)
  )
```
:::

#### Part C: Check your work

::: {.callout-warning appearance="minimal"}
##### Task

Use `tabyl()` to compare `degree` and `degree_filled` in `hrs_12_degree_filled` (hint: `tabyl()` can take two column names to cross-tabulate them). Confirm that every row that was `NA` in `degree` now shows `"Unknown"` in `degree_filled`.
:::

::: blue
**Answer:**

Not given
:::

#### Part D: Turn a valid value into `NA`

::: {.callout-warning appearance="minimal"}
##### Task

Suppose, just for practice, we wanted to treat everyone who answered `"No"` to `drink` (whether they drink alcohol) as missing instead (this is **not** something you'd actually want to do with real data, but it shows how the function works). First, use `mutate()` to make a copy of the column called `drink_na`. Then use `replace_with_na()` to replace `"No"` with `NA` in the `drink_na` column. Assign the result to `hrs_13_replace_with_na`.

Use `tabyl()` to compare the original `drink` column to your new `drink_na` column and confirm the replacement worked as expected.
:::

::: blue
**Answer:**

Start with something like:

```{r}
#| eval: false

hrs_13_replace_with_na <- hrs_12_degree_filled |> 
  mutate(drink_na = drink) |> 
  replace_with_na(replace = list(
    ______ = ______
  ))
```
:::

#### Part E: Summarize your final, cleaned dataset

::: {.callout-warning appearance="minimal"}
##### Task

Take `hrs_12_degree_filled` (from Part B, before Part D's practice-only recoding) and run `miss_var_summary()` and `gt()` on it one more time. Assign it to `hrs_14_summary_after`. Compare it to `hrs_06_summary_before` from Question 3 Part A. Which variables now show less missingness, and why?
:::

::: blue
**Answer:**

Not given
:::
