---
title: "Practice 6 Answers"
subtitle: "PUBH 523/623"
author: "Nicky Wakim"
date-modified: "today"
format: 
  html:
    link-external-newwindow: true
    toc: true
    self-contained: true
    page-layout: full
    html-math-method: mathjax
    self-contained-math: true
    highlight-style: ayu
    code-block-bg: "#f8f9fa"   
    code-block-border-left: "#5BAFF8" 
editor_options: 
  chunk_output_type: console
---

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#| warning: false

pacman::p_load(
  tidyverse, 
  rio, 
  here,
  skimr,
  janitor,
  rstatix,
  gtsummary,
  gt
)

hrs_data <- import(here("data", "hrs_data.rds")) #<1>
```

1. This file path is specific to my folder set-up.

## Questions

### Question 1: Data Summarization: Intro

In Lesson 24, we saw that there are many ways to summarize data in R, and that the right tool often depends on whether a variable is categorical or numeric.

#### Part A: Categorical vs. numeric

::: {.callout-warning appearance="minimal"}
##### Task

In 1-2 sentences, explain why we might need different summary tools for categorical data versus numeric data. What does each type of tool tend to focus on?
:::

::: blue
**Answer:**

Not given
:::

#### Part B: Sort the functions

::: {.callout-warning appearance="minimal"}
##### Task

For each function below, fill in the table with:

1.  Which package it comes from
2.  Whether it works on **categorical**, **numeric**, or **both** types of data
3.  What you'd use it for (a quick browse of a whole dataset, a custom/pointed summary, a grouped summary, or a presentation-ready table — more than one may apply)
4.  What it does, in your own words

| Function | Package | Categorical, Numeric, or Both? | What would you use it for? | What does it do? |
|--------------|-----------|-----------|--------------------------|----------------|
| `summary()` |  |  |  |  |
| `skim()` |  |  |  |  |
| `get_summary_stats()` |  |  |  |  |
| `summarise()` |  |  |  |  |
| `tabyl()` |  |  |  |  |
| `tbl_summary()` |  |  |  |  |
:::

::: blue
**Answer:**

Example of `tabyl()`:

| Function | Package | Categorical, Numeric, or Both? | What would you use it for? | What does it do? |
|----|----|----|----|----|
| `summary()` |  |  |  |  |
| `skim()` |  |  |  |  |
| `get_summary_stats()` |  |  |  |  |
| `summarise()` |  |  |  |  |
| `tabyl()` | `janitor` | Categorical | A custom/pointed summary, a presentation-ready table | Builds a frequency table (or cross-tabulation) of one or more categorical variables |
| `tbl_summary()` |  |  |  |  |
:::

### Question 2: Data Summarization: Quick summaries

In Lesson 25, we covered three functions for quickly browsing a dataset: `summary()`, `skim()`, and `get_summary_stats()`.

#### Part A: `summary()`

::: {.callout-warning appearance="minimal"}
##### Task 1

Use `summary()` on all of `hrs_data`. Look through the output but do not assign it to anything. Do you notice anything interesting about the summary statistics of `HHID` (our household identifier) and `id` (our observation identifier)?
:::

::: blue
**Answer:**

Here's an example to show you the pattern — the rest of your answers should follow this same style, with code chunks and a `#| ...` chunk options as needed.

```{r}
summary(hrs_data)
```
:::

::: {.callout-warning appearance="minimal"}
##### Task 2

Now use `summary()` on just `hrs_data$income`. What is the median income?
:::

::: {.callout-tip appearance="minimal"}
##### Tip

`summary()` treats categorical (factor) and numeric columns differently — categorical columns show counts per category, while numeric columns show min/median/mean/max, etc.
:::

::: blue
**Answer:**

Not given
:::

#### Part B: `skim()`

::: {.callout-warning appearance="minimal"}
##### Task 1

Use `skim()` on all of `hrs_data`. What variables have missing categories from the `top_counts`?
:::

::: blue
**Answer:**

Not given
:::

::: {.callout-warning appearance="minimal"}
##### Task 2

Using `select()`, first keep only `cesd`, `cond_count`, and `income` from `hrs_data`, then pipe the result into `skim()`. Based on the output, which of these three variables has the most missing values, and how many?
:::

::: blue
**Answer:**

Not given
:::

#### Part C: `get_summary_stats()`

::: {.callout-warning appearance="minimal"}
##### Task 1

Use `get_summary_stats()` on `hrs_data` to get the `"common"` summary statistics for every numeric variable. Assign the result to `hrs_01_summary_stats`.

What is the IQR for `ed`?
:::

::: blue
**Answer:**

Start with something like:

```{r}
#| eval: false

hrs_01_summary_stats <- hrs_data |> 
  get_summary_stats(type = ______)
```
:::

::: {.callout-warning appearance="minimal"}
##### Task 2

Now use `get_summary_stats()` again, but this time restrict it to just `age_yr`, `cesd`, and `income`, and set `type = "mean_sd"`. Assign the result to `hrs_02_summary_stats_meansd`.

What is the standard deviation of `cesd`?
:::

::: blue
**Answer:**

Not given
:::

::: {.callout-tip appearance="minimal"}
##### Tip

`get_summary_stats()` only works on numeric variables — any categorical variables you feed it will be silently dropped.
:::

### Question 3: Data Summarization: Tables

In Lesson 26, we covered two functions for building summary tables: `tabyl()` and `tbl_summary()`, plus `gt()` for presentation.

#### Part A: `tabyl()`

::: {.callout-warning appearance="minimal"}
##### Task 1

Use `tabyl()` to build a frequency table of `srh` (self-rated health). Assign the result to `hrs_03_tabyl_srh`.
:::

::: blue
**Answer:**

Example:

```{r}
hrs_data |> 
  tabyl(srh)
```
:::

::: {.callout-warning appearance="minimal"}
##### Task 2

Use `tabyl()` to cross-tabulate `srh` by `sleep`. Assign the result to `hrs_04_tabyl_cross`.
:::

::: blue
**Answer:**

Not given
:::

#### Part B: `adorn_*()`

::: {.callout-warning appearance="minimal"}
##### Task

Starting from `hrs_04_tabyl_cross`, add a totals row, convert the counts to column percentages, format the percentages to 1 decimal place, and add the raw counts back in front of the percentages. Assign the result to `hrs_05_tabyl_formatted` and display the table.
:::

::: {.callout-tip appearance="minimal"}
##### Tip

Order matters when chaining `adorn_*()` functions together: tabulate → add totals → convert to percentages → format as percents → add counts back.
:::

::: blue
**Answer:**

Start with something like:

```{r}
#| eval: false

hrs_05_tabyl_formatted <- hrs_04_tabyl_cross |> 
  adorn_totals(______) |> 
  adorn_percentages(______) |> 
  adorn_pct_formatting(______) |> 
  adorn_ns(______)
```
:::

#### Part C: `tbl_summary()`

::: {.callout-warning appearance="minimal"}
##### Task 1

Using `select()`, keep only `age_yr`, `act_vig`, `bp`, `lung`, `cesd`, and `height` from `hrs_data`, then pipe the result into `tbl_summary()`. Assign the result to `hrs_06_tbl_summary`.

- How many missing values are there for "Ever Had High Blood Pressure"?
- Do you notice any difference between the summary statistics for `cesd` using `tbl_summary()` and `skim()`?
:::

::: blue
**Answer:**

Not given
:::

::: {.callout-warning appearance="minimal"}
##### Task 2

Starting over from `hrs_data`, build another `tbl_summary()` with just `age_yr` and `height`, but this time show `{mean} ({sd})` for both variables instead of the defaults, and give `height` the display label `"Self-Reported Height (m)"` to make sure readers know height is measured in meters. Assign the result to `hrs_07_tbl_summary_custom` and display the table.
:::

::: {.callout-tip appearance="minimal"}
##### Tip

Since both variables are numeric, we can use the following input to make ALL numeric variables show mean and standard deviation: 

```{r}
#| eval: false
`statistic = all_continuous() ~ "{mean} ({sd})"`
```

:::

::: blue
**Answer:**

Start with something like:

```{r}
#| eval: false

hrs_07_tbl_summary_custom <- hrs_data |> 
  select(______) |> 
  tbl_summary(
    statistic = ______,
    label = height ~ ______
  )
```
:::

#### Part D: `gt()`

::: {.callout-warning appearance="minimal"}
##### Task

Take `hrs_03_tabyl_srh` from Part A and pipe it into `gt()`, then use `tab_header()` to give the table the title `"Self-Rated Health"`. Assign the result to `hrs_08_gt`.
:::

::: blue
**Answer:**

Not given
:::

### Question 4: Data Summarization: Grouped summaries

In Lesson 27, we covered how to compare summary statistics across groups using `group_by()` with `summarise()`, `skim()`, and `get_summary_stats()`, as well as the `by =` argument in `tbl_summary()`.

#### Part A: `summarise()` with `group_by()`

::: {.callout-warning appearance="minimal"}
##### Task

Group `hrs_data` by `degree`, then use `summarise()` to calculate the mean and standard deviation of `income` for each group (remember to handle missing values). Assign the result to `hrs_09_grouped_summarise`.

Which `degree` group has the highest mean income?
:::

::: blue
**Answer:**

The resulting grouped summary would look something like this:

```{r}
#| echo: false

hrs_09_grouped_summarise <- hrs_data |> 
  group_by(degree) |> 
  summarise(
    mean_income = mean(income, na.rm = TRUE),
    sd_income = sd(income, na.rm = TRUE)
  )

hrs_09_grouped_summarise
```
:::

#### Part B: `skim()` grouped

::: {.callout-warning appearance="minimal"}
##### Task

Group `hrs_data` by `sex`, use `select()` to keep only `cesd` and `cond_count`, and then pipe the result into `skim()`. Look through the output but do not assign it to anything.

Does the average `cesd` score differ between males and females?
:::

::: blue
**Answer:**

Not given
:::

#### Part C: `get_summary_stats()` grouped

::: {.callout-warning appearance="minimal"}
##### Task

Group `hrs_data` by `bp` (high blood pressure), then use `get_summary_stats()` on `age_yr` and `cond_count` with `type = "common"`. Assign the result to `hrs_10_grouped_stats`.

How does mean `age_yr` compare between those with and without high blood pressure?
:::

::: blue
**Answer:**

Not given
:::

#### Part D: `tbl_summary()` with `by =`

::: {.callout-warning appearance="minimal"}
##### Task

Using `select()`, keep `age_yr`, `cesd`, `srh`, and `diab` from `hrs_data`, then use `tbl_summary()` to stratify the table by diabetes status. Assign the result to `hrs_11_tbl_summary_grouped` and display the table. 
:::

::: blue
**Answer:**

Not given
:::
