---
title: "Practice 5 Answers"
subtitle: "PUBH 523/623"
author: "Nicky Wakim"
date-modified: "today"
format: 
  html:
    link-external-newwindow: true
    toc: true
    self-contained: true
    page-layout: full
    html-math-method: mathjax
    self-contained-math: true
    highlight-style: ayu
    code-block-bg: "#f8f9fa"   
    code-block-border-left: "#5BAFF8" 
editor_options: 
  chunk_output_type: console
---

```{=html}
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#| warning: false

pacman::p_load(
  tidyverse, 
  rio, 
  here
)

hrs_data <- import(here("data", "hrs_data.rds")) #<1>
```

1. This file path is specific to my folder set-up. 

## Questions

### Question 1: `dplyr` Basics

In Lesson 20, we learned that `dplyr` functions are "verbs" that take a data frame in and return a data frame out, and that they can be grouped into four categories based on what they operate on: rows, columns, groups, and joins.

#### Part A: Input and output

::: {.callout-warning appearance="minimal"}
##### Task

In 1 sentence, describe the general input and output of most `dplyr` functions. What type of object typically goes in, and what type of object comes out?
:::

::: blue
**Answer:**

Not given
:::

#### Part B: Sort the verbs

::: {.callout-warning appearance="minimal"}
##### Task

For each function below, fill in the table with:

1. Whether it operates on **rows**, **columns**, or is part of the **groups** category (which we have *not* covered yet, but you saw it mentioned in Lesson 20)
2. What the function does, in your own words
3. What output you'd expect if you ran it on `hrs_data` (for example: the same number of rows but fewer columns, fewer rows but all the same columns, the rows reordered, etc.)

| Function | Rows, Columns, or Groups (not yet covered)? | What does it do? | Expected output on `hrs_data` |
|----------|----------------------------------------------|-------------------|--------------------------------|
| `filter()` |  |  |  |
| `select()` |  |  |  |
| `mutate()` |  |  |  |
| `arrange()` |  |  |  |
| `distinct()` |  |  |  |
| `relocate()` |  |  |  |
| `rename()` |  |  |  |
| `group_by()` |  |  |  |
| `summarize()` |  |  |  |
:::

::: blue
**Answer:**

Example of mutate:

| Function | Rows, Columns, or Groups (not yet covered)? | What does it do? | Expected output on `hrs_data` |
|----------|----------------------------------------------|-------------------|--------------------------------|
| `filter()` |  |  |  |
| `select()` |  |  |  |
| `mutate()` | Columns | Creates a new column, or changes the values in an existing one | Same rows, same original columns plus any new column(s) added |
| `arrange()` |  |  |  |
| `distinct()` |  |  |  |
| `relocate()` |  |  |  |
| `rename()` |  |  |  |
| `group_by()` |  |  |  |
| `summarize()` |  |  |  |
:::

### Question 2: Data transformation: rows

In Lesson 21, we learned three functions for transforming rows: `filter()`, `arrange()`, and `distinct()`.

#### Part A: `filter()`

Using `hrs_data`:

::: {.callout-warning appearance="minimal"}
##### Task 1

Use `filter()` to keep only respondents whose `cond_count` (total number of chronic conditions) is **5 or more**. Assign this to `hrs_01_filter_conditions`, and use `nrow()` to report how many respondents that is.
:::

::: blue
**Answer:**

Here's an example to show you the pattern — the rest of your answers should follow this same style, with code chunks and a `#| ...` chunk options as needed.

```{r}
hrs_01_filter_conditions <- hrs_data |> 
  filter(cond_count >= 5)

nrow(hrs_01_filter_conditions)
tibble(hrs_01_filter_conditions)
```
:::

::: {.callout-warning appearance="minimal"}
##### Task 2

Use `filter()` to keep only respondents whose `srh` (self-rated health) is exactly `"Poor"`. Assign this to `hrs_02_filter_srh`.
:::

::: {.callout-tip appearance="minimal"}
##### Tip

Remember: if you misspell or mis-capitalize a value for a categorical/factor variable, `filter()` will **not** throw an error — it will just silently return 0 rows. Always check your output!
:::

::: blue
**Answer:**

Not given
:::

#### Part B: `arrange()`

::: {.callout-warning appearance="minimal"}
##### Task 1

Sort `hrs_data` by `cesd` in **descending** order first, and within ties, by `age_yr` in **ascending** order. Assign the result to `hrs_03_arrange`, and print it using `tibble()`.
:::

::: blue
**Answer:**

Not given
:::

::: {.callout-warning appearance="minimal"}
##### Task 2

In 1 sentence, describe what would change about the output if you arranged by `age_yr` first and `cesd` second instead.
:::

::: blue
**Answer:**

Not given
:::

#### Part C: `distinct()`

::: {.callout-warning appearance="minimal"}
##### Task 1

Use `distinct()` on `degree` and `race_original` together to see how many unique combinations of these two variables exist in `hrs_data`. Assign this to `hrs_04_distinct`, and print it using `tibble()`.
:::

::: blue
**Answer:**

The resulting distinct combinations would look like this:

```{r}
#| echo: false

hrs_04_distinct <- hrs_data |> 
  distinct(degree, race_original)

hrs_04_distinct
```
:::

::: {.callout-warning appearance="minimal"}
##### Task 2

Use `distinct()` on `degree` again, but this time with `.keep_all = TRUE`, so you keep the full row (all columns) for the first respondent found at each unique `degree` value. Assign this to `hrs_05_distinct_all`.
:::

::: blue
**Answer:**

Not given
:::

### Question 3: Data transformation: columns

In Lesson 22, we learned four functions for transforming columns: `select()`, `relocate()`, `mutate()`, and `rename()`.

#### Part A: `select()` and `relocate()`

::: {.callout-warning appearance="minimal"}
##### Task 1

Use `select()` to keep only `id`, `age_yr`, `ed`, `degree`, `income`, `cesd`, and `srh` from `hrs_data`. Assign this to `hrs_06_vars_subset`.
:::

::: blue
**Answer:**

Not given
:::

::: {.callout-warning appearance="minimal"}
##### Task 2

Starting from `hrs_06_vars_subset`, use `relocate()` to move `srh` so that it comes right after `id`. Assign this to `hrs_07_reloc`, and print it using `tibble()`.
:::

::: blue
**Answer:**

Not given
:::

#### Part B: `mutate()`

::: {.callout-warning appearance="minimal"}
##### Task

Using `hrs_data`, create a new variable called `height_ft` that converts `height` (recorded in meters) into feet, using the conversion 1 meter = 3.28084 feet. Round `height_ft` to 1 decimal place using `round()` *inside* your `mutate()` call. Assign the result to `hrs_08_mutate`, and use `relocate()` so that `height_ft` sits right next to `height` for easy comparison. Print it using `tibble()`.
:::

::: {.callout-tip appearance="minimal"}
##### Tip

Remember, inside `mutate()` you use a single `=`, not `<-` or `==`, to define your new column.
:::

::: blue
**Answer:**

Start with something like:

```{r}
#| eval: false

hrs_08_mutate <- hrs_data |> 
  mutate(height_ft = ______ )
```
:::

#### Part C: `rename()`

::: {.callout-warning appearance="minimal"}
##### Task

Using `hrs_data`, rename `cesd` to `depression_score` and `srh` to `self_rated_health`. Assign the result to `hrs_09_rename`, and print `names(hrs_09_rename)` to confirm the change.
:::

::: blue
**Answer:**

Not given
:::

### Question 4: Data transformation: advanced

In Lesson 23, we covered more advanced techniques: combining conditions with `&`, `|`, and `%in%`; checking our work; creating variables with `case_when()`; and working with factors.

#### Part A: Combining conditions

::: {.callout-warning appearance="minimal"}
##### Task 1

Before writing any code: do you expect *more* respondents to have diabetes **AND** high blood pressure (`bp`), or diabetes **OR** high blood pressure? Explain your reasoning in 1 sentence.
:::

::: blue
**Answer:**

Not given
:::

::: {.callout-warning appearance="minimal"}
##### Task 2

Use `filter()` with `&` to find respondents who **have diabetes and have high blood pressure**. Assign this to `hrs_10_filter_diab_bp`.
:::

::: blue
**Answer:**

Start with something like:

```{r}
#| eval: false

hrs_10_filter_diab_bp <- hrs_data |> 
  filter(______ & _______)
```
:::

::: {.callout-warning appearance="minimal"}
##### Task 3

Use `filter()` with `|` to find respondents who **have diabetes or have high blood pressure**. Assign this to `hrs_11_filter_diab_or_bp`.
:::

::: blue
**Answer:**

Not given
:::

::: {.callout-warning appearance="minimal"}
##### Task 4

Print `hrs_10_filter_diab_bp` and `hrs_11_filter_diab_or_bp` using `tibble()`, and reference the row count shown at the top of each output to compare how many respondents are in each data frame. Then use `distinct()` on `diab` and `bp` within each object to see which combinations of `diab` and `bp` are actually present in each dataset. Does this match your prediction from Task 1?
:::

::: blue
**Answer:**

Not given
:::

::: {.callout-warning appearance="minimal"}
##### Task 5

Now combine **multiple** logic statements in a single `filter()`: find respondents who **have diabetes and have high blood pressure, or who have more than 5 chronic conditions**. Assign this to `hrs_12_filter_combo`. Think carefully about where your parentheses go.
:::

::: blue
**Answer:**

Not given
:::

#### Part B: `case_when()`

::: {.callout-warning appearance="minimal"}
##### Task

Using `hrs_data`, create a new variable called `income_cat` with **4** categories based on `income`:

- `"<25k"` if `income` is less than \$25,000
- `"25k-50k"` if `income` is at least \$25,000 but less than \$50,000
- `"50k+"` if `income` is \$50,000 or more
- `NA` if `income` is missing (use `.default` for this case)

Then use `fct()` to convert `income_cat` into a factor. Assign your final result to `hrs_13_case_when`, and check your work using `count()`.
:::

::: {.callout-tip appearance="minimal"}
##### Tip

For the middle category, you'll need an `&` statement inside `case_when()`, similar to how we combined conditions in Part A.
:::

::: blue
**Answer:**

Start with something like: 

```{r}
#| eval: false

hrs_13_case_when <- hrs_data |> 
  mutate(
    income_cat = case_when(
      ______ ~ ______,
      ______ ~ ______,
      ______ ~ ______,
      .default = ______
    )
  )
```
:::

#### Part C: Factors

::: {.callout-warning appearance="minimal"}
##### Task 1

Check the current class of `hrs_data$srh` using `class()`.
:::

::: blue
**Answer:**

Example:

```{r}
class(hrs_data$srh)
```
:::

::: {.callout-warning appearance="minimal"}
##### Task 2

Use `fct_relevel()` to reorder the levels of `srh` into descending order from "Excellent" to "Poor." Call the new column `srh_releveled`, and assign the result to `hrs_14_relevel`. Confirm the new order with `levels()`.
:::

::: {.callout-tip appearance="minimal"}
##### Tip

If you misspell one of the levels inside `fct_relevel()`, R will not throw an error — it will silently drop that level to the bottom of the order (and print a warning in the console). Double-check your spelling against the original values in `srh`!
:::

::: blue
**Answer:**

Not given
:::

::: {.callout-warning appearance="minimal"}
##### Task 3

Now suppose you had instead written `fct_relevel(srh, "Excellent", "Very good", "Good", "Fair", "Poor")` — with a lowercase "g" in `"Very good"`. Run this version and look at the warning message R gives you. In 1-2 sentences, explain what the warning is telling you and why it happens.
:::

::: blue
**Answer:**

Not given
:::

#### Part D: Selecting columns by pattern or type

::: {.callout-warning appearance="minimal"}
##### Task 1

Use `select()` with `contains()` to grab all columns in `hrs_data` whose name contains `"smoke"`. Assign this to `hrs_15_select_contains`, and print it using `tibble()`.
:::

::: blue
**Answer:**

Start with something like:

```{r}
#| eval: false

hrs_15_select_contains <- hrs_data |> 
  select(________)
```
:::

::: {.callout-warning appearance="minimal"}
##### Task 2

Use `select()` with `where(is.factor)` to grab all factor columns in `hrs_data`, and pipe the result into `names()` so you just see the column names.
:::

::: blue
**Answer:**

Not given
:::
