---
title: "Practice 3"
subtitle: "PUBH 523/623"
author: "Your name here"
date-modified: "today"
format: 
  html:
    link-external-newwindow: true
    toc: true
    self-contained: true
    page-layout: full
    html-math-method: mathjax
    self-contained-math: true
    highlight-style: ayu
    code-block-bg: "#f8f9fa"   
    code-block-border-left: "#5BAFF8" 
editor_options: 
  chunk_output_type: console
---

## Overview

This practice is mostly **hands-on**. You will get familiar with the HRS dataset we will use for the rest of the course, practice calling functions and reading their help files, work with different types of R objects, and build up your skills for spotting and fixing common errors — all skills you will rely on every week.

Topics covered (Lessons 11-15):

- Introduction to HRS dataset
- Functions
- Objects
- Common errors
- Getting help

## Directions

Download this practice `.qmd` from the course GitHub repository, rename it, and save it in your course folder.

1. [Download `practice_03.qmd` from GitHub.](https://github.com/nwakim/PUBH_523_26Su/blob/main/practice/practice_03.qmd)
2. Rename the file to `Lastname_Firstinitial_Practice_03.qmd` (or `lastname_firstinitial_practice_03.qmd`), replacing with your actual last name and first initial.
3. Save it inside your `practice` folder. You may create a `practice_04` subfolder if you prefer.

::: {.callout-tip appearance="minimal"}
##### How to use this file

- The boxes with the **yellow stripe** explain what to do. Keep these in your submitted file.
- Do all your work **below and outside** of the yellow-striped boxes. 
  - *You will need to create code chunks within this document to run your R code.*
- Please delete the following sections before submitting:
    - Overview
    - Directions
    - Tip boxes (green stripe)
    - Caution boxes / Needed Steps (orange/red stripe)
    - Submission checklist
    - Note on AI usage
:::

## Questions

### Question 1: Getting to Know the HRS Dataset

In Lesson 11, we were introduced to the *synthetic* HRS (Health and Retirement Study) dataset we will use for the rest of the course.

#### Part A: Explore the codebook

::: {.callout-warning appearance="minimal"}
##### Task

Open the [HRS codebook](https://nwakim.github.io/PUBH_523_26Su/resources/HRS_codebook.html) and answer the following:

1. What does the variable `srh` measure, and what are its possible levels?
2. Name **two** variables in the dataset that are stored as `factor`, and **two** that are stored as `numeric`.
3. In 1–2 sentences, explain why this is a *synthetic* dataset rather than the real HRS data, and what that means for any patterns you might see in it.
:::

*Write your answers here.*

#### Part B: A mini HRS sample

::: {.callout-tip appearance="minimal"}
##### Note

We have not yet learned how to import outside data files into R (that's coming in a future lesson!). For now, here is a small, made-up sample of 5 "respondents" with a few variables from the codebook, built directly in R. Run the chunk below so `mini_hrs` is available for the rest of this practice — you don't need to change anything in it.
:::

```{r}
#| message: false
pacman::p_load(tidyverse)

mini_hrs <- data.frame(
  id     = c("R1", "R2", "R3", "R4", "R5"),
  age_yr = c(55, 62, 71, 48, 67),
  sex    = c("Female", "Male", "Male", "Female", "Male"),
  cesd   = c(2, 4, 6, 0, NA)
)

mini_hrs
```

::: {.callout-warning appearance="minimal"}
##### Task

Compare `mini_hrs` to the [HRS codebook](https://nwakim.github.io/PUBH_523_26Su/resources/HRS_codebook.html). In 2–3 sentences, describe what one row of `mini_hrs` represents, and name one way this mini sample is similar to (or different from) the full synthetic HRS dataset described in Lesson 11.
:::

*Write your answer here.*

### Question 2: Functions

In Lesson 12, we learned that functions take the general form `function_name(argument1 = value1, argument2 = value2, ...)`.

#### Part A: Call a function with arguments

::: {.callout-warning appearance="minimal"}
##### Task

Use the `seq()` function to generate a sequence of even numbers from 2 to 20, naming your arguments (`from`, `to`, `by`) explicitly. Assign the output to an object called `evens`, then print `evens`.
:::

#### Part B: Use a function on your data

::: {.callout-warning appearance="minimal"}
##### Task

Using your `mini_hrs` data frame from Question 1, calculate the mean of the `cesd` column using the `mean()` function. Access the column with the `$` operator (Lesson 13).
:::

::: {.callout-tip appearance="minimal"}
##### Tip

Look closely at the `cesd` column in `mini_hrs` — one value is missing (`NA`). You'll need to look at the help file for `mean()` to figure out which argument handles missing values, or your result will print as `NA`.
:::

#### Part C: Look up a function's help file

::: {.callout-warning appearance="minimal"}
##### Task

Pick any **one** function you used in Question 1 or 2 (for example, `data.frame()`, `seq()`, or `mean()`). Open its help file using `?`. In 2–3 sentences, describe what you found: what are the function's main arguments, and is there anything in the help file that surprised you or that you didn't know before?
:::

*Write your answer here.*

### Question 3: Objects

In Lesson 13, we learned about different types of R objects, both single pieces of data and collections of data.

#### Part A: Object types

::: {.callout-warning appearance="minimal"}
##### Task

Fill in the table below with an example value for each object class, and one sentence describing when you might use that type in the HRS dataset.
:::

| Class | Example value | When might you see this in the real HRS dataset (see codebook)? |
|-------|----------------|------------------------------------------|
| Double (`dbl`) |  |  |
| Character (`chr`) |  |  |
| Factor (`fct`) |  |  |
| Logical (`lgl`) |  |  |

#### Part B: Check the class of variables

::: {.callout-warning appearance="minimal"}
##### Task

Using your `mini_hrs` data frame, use the `class()` function to check the object type of each of its four columns. Access each column with the `$` operator.
:::

*For each column, state whether the class is what you would have expected before running `class()`.*

### Question 4: Common Errors

Even experienced R users run into errors constantly! Lesson 14 covered some of the most common ones.

#### Part A: Spot and fix the bugs

::: {.callout-warning appearance="minimal"}
##### Task

Each code chunk below contains **one error**. For each chunk, (1) identify what is causing the error, and (2) fix the code so it runs correctly.
:::

```{r}
#| eval: false
# Chunk 1
mean(mini_hrs$Cesd)
```

*What's wrong with Chunk 1?*

```{r}
#| eval: false
# Chunk 2
avg_age <- mean(mini_hrs$age_yr
```

*What's wrong with Chunk 2?*

```{r}
#| eval: false
# Chunk 3
ages <- c(55, 62, 71, 48, 67)
avg <- mean("ages")
```

*What's wrong with Chunk 3?*

::: {.callout-tip appearance="minimal"}
##### Tip

Common culprits include: capitalization/spelling of variable or object names, unmatched parentheses, missing commas, quoting something that shouldn't be quoted (or vice versa), and forgetting to load a package.
:::

#### Part B: Read an error message

::: {.callout-warning appearance="minimal"}
##### Task

We ran the following code chunk and got the printed error. In 1–2 sentences, explain what the error message is telling you.
:::

```{r}
#| error: true
mini_hrs$age_yr + mini_hrs$id
```

*Explain what the error message means:*

### Question 5: Getting Help

Knowing how to get *unstuck* is one of the most important skills in R. Lesson 15 covered strategies for getting help.

#### Part A: Search for your error

::: {.callout-warning appearance="minimal"}
##### Task

Take the error message from Question 4 Part B. Search for it online (Google, Stack Overflow, or another resource). Link to a resource you found helpful, and in 1–2 sentences summarize what it suggested.
:::

*Link and summary here:*

#### Part B: Ask a good question

::: {.callout-warning appearance="minimal"}
##### Task

Imagine you still can't solve the error from Question 4 Part B, and you need to email your instructor for help. Write the email you would send. Your email should include the pieces of information a good help-seeking question needs (think about what you'd want to know if you were the one answering).
:::

*Write your email here.*

::: {.callout-tip appearance="minimal"}
##### Tip

Refer back to Lesson 15 for what makes a question easy for someone else to help with — for example, showing your actual code and the exact error message, rather than just saying "it doesn't work."
:::

## Submission Checklist

If having a checklist helps you stay organized, check off each item below as you complete it. You do not need to submit this checklist.

- [ ] Codebook questions answered (Q1 Part A)
- [ ] `mini_hrs` data frame run and compared to codebook (Q1 Part B)
- [ ] `evens` sequence created with `seq()` (Q2 Part A)
- [ ] Mean of `cesd` calculated, with explanation of any argument added (Q2 Part B)
- [ ] Help file explored for one function, with summary written (Q2 Part C)
- [ ] Object type table filled in (Q3 Part A)
- [ ] Class of each column in `mini_hrs` checked with `class()` (Q3 Part B)
- [ ] Data frame vs. tibble question answered (Q3 Part C)
- [ ] All three bugs identified and fixed (Q4 Part A)
- [ ] Error message pasted and explained (Q4 Part B)
- [ ] Online resource for the error found and summarized (Q5 Part A)
- [ ] Help-seeking email drafted (Q5 Part B)
- [ ] Document renders to HTML with no errors
- [ ] Files named correctly: `Lastname_Firstinitial_Practice_03.qmd` and `Lastname_Firstinitial_Practice_03.html` (or lowercase versions)
- [ ] Both `.qmd` and `.html` submitted to Brightspace

## Note on AI usage

I used GenAI (Claude) to help me draft this practice assignment. It helped me brainstorm ideas for the assignment, and I directed it to help you complete the needed tasks that I showed in the lessons.