---
title: "Data Summarization: Intro"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Data Summarization: Intro"
    html-math-method: mathjax
    highlight-style: ayu
    code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

pacman::p_load(
  here, 
  tidyverse, 
  writexl, 
  rstatix, 
  janitor, 
  naniar,
  gt,
  gtsummary,
  readxl, 
  rio
)
```

```{r}
#| label: "import data" 
#| echo: false
hrs_data_rds <- import(here("data", "hrs_data.rds"))
hrs_00 <- clean_names(hrs_data_rds) |>
  relocate(diab, degree, .after = id)
hrs_degree_chr = hrs_00 |>
  mutate(degree = as.character(degree))
  # select(id, age_yr, ed, cesd, degree, srh)
```

## Data Summarization

- There are **many ways to summarize data** in R
  - Each function can be useful in different scenarios!

 
  
- We may need **different tools based on the type of measurement** (categorical vs. numeric)
  - Categorical data (represented as factors in R) will focus on frequencies and counts
  - Numeric data (represented as numbers in R) will focus on summary metrics like mean, median, standard deviation, etc.

## Summary functions we will cover

| Package | Function | Categorical or Numeric? | Type | What it does |
|--------|-----------|-----------|-----------------|---------------------|
| built-in | `summary()` | Both | Quick browsing | Base R's simple way to get a quick stats of each variable |
| `skimr` | `skim()` | Both | Quick browsing, Grouped summaries | Overview of every variable in your dataset at once |
| `rstatix` | `get_summary` `_stats()` | Numeric | Quick browsing, Grouped summaries | Automatically generates common numeric summary statistics |
| `dplyr` | `summarize()` | Both | Pointed summaries, Grouped summaries | Calculates custom summary statistics you specify yourself |
| `janitor` | `tabyl()` | Categorical | Quick browsing | Builds frequency tables for one or more categorical variables |
| `gtsummary` | `tbl_summary()` | Both | Presentations, Grouped summaries | Creates a polished, publication-ready summary table |

## Helper functions we will cover

| Package | Function | Type | What it does |
|-----------|-----------|---------------|----------------------|
| `dplyr` | `group_by()` | Grouped summaries | Splits your data into groups so later functions operate *per group* |
| `gt` | `gt()` | Presentation | Creates a polished, publication-ready table |

## Make sure you have the following packages installed

- Remember, you can use `pacman::p_load()` to install and load packages in one step

- Needed packages:

  - `dplyr`
  - `skimr`
  - `rstatix`
  - `janitor`
  - `gtsummary`
  - `gt`
