---
title: "Missing data: transformations"
author: "Nicky Wakim, Jessica Minnier, and Meike Niederhausen"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Missing data: transformations"
    html-math-method: mathjax
    highlight-style: ayu
    code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: setup
#| include: false

pacman::p_load(
  tidyverse, 
  here, 
  rio, 
  naniar, 
  janitor,
  gt
)

hrs_data <- import(here("data", "hrs_data.rds")) |>
  mutate(srh = fct_relevel(srh, "Excellent", "Very Good", "Good", "Fair", "Poor"))
```

# From finding missing data to handling it {visibility="hidden"}

## Before we start

- Install and load the `naniar` package (a play on Narnia) if you haven't already!

![[Artwork by @allison_horst](https://allisonhorst.com/r-packages-functions)](../img_slides/horst_naniar.png){fig-alt="\"Explore missing data in naniar!\" Cute monsters with exploring gear looking around a landscape full of subtle (and not so subtle) NAs, hidden in trees, on mountain tops, as birds, and in flowers. Learn more about naniar."}


- We'll be working with only a few variables in the HRS dataset in this lesson!
  - Mostly to help us view the tables and plots within the slides
  
```{r}
hrs_00 <- hrs_data |>
  select(ed, race_original, srh, sex, act_vig, height, degree)
```


## What do we do with missing data?

- *Filtering out* missing values with `drop_na()`
- A common gotcha: how `filter()` silently drops `NA`s

 

- We will also use `naniar` functions to work with `NA`s in a more flexible way
  - *Filling in* missing values with `replace_na()` (and `fct_na_value_to_level()` for factors)
  - *Replacing* a values (numbers or levels) with `NA` using `replace_with_na()`
  
 

- Not covered in this course (but very important in practice): 
  - Imputation methods for filling in missing values
  - Weighting methods for adjusting for missingness


# Filtering out missing values using `drop_na()` {visibility="hidden"}

## Check missingness rates first before dropping `NA`s

- We may want to remove rows that have missing data, coded as `NA`

- First, summarize and visualize missingness using `naniar`

- We'll take a quick look at the counts

```{r}
#| code-fold: true
#| code-summary: "Create missingness sumamry table with `miss_var_summary()`"
hrs_00 |>
  miss_var_summary() |>
  gt() |>
  tab_options(table.font.size = px(40))
```

## `drop_na()`: removing missingness from all variables in the dataset

- `drop_na()` removes rows with missing data
  - You can target specific columns, or drop any row with any missingness

```{r}
hrs_01_na_rm <- hrs_00 |>
  drop_na()
tibble(hrs_01_na_rm)
```

*Pay attention to the number of rows for each of these results*

## `drop_na()`: removing missingness from all variables in the dataset

- Remove missingness in one or more columns:

```{r}
hrs_02_race_rm <- hrs_00 |> 
  drop_na(race_original, ed)

hrs_02_race_rm |> miss_var_summary() |>
  gt() |>
  tab_options(table.font.size = px(40))
```
  

# A warning about `filter()`: check for `NA` {visibility="hidden"}

## Warning on `filter()`: it can silently drop your `NA`s

- Suppose we want to remove the `"Poor"` category from `srh` (self-reported health)

::: columns
::: column
```{r}
hrs_00 |> tabyl(srh)
```
:::
::: column


`srh` has 4 missing values!
:::
:::

::: columns
::: column
When we try to filter using `!=`, we **lose our `NA`s**!

```{r}
hrs_03_rm_poor <- hrs_00 |> 
  filter(srh != "Poor")
```
:::
::: column
When `R` checks `NA != "Poor"`, it evaluates to `NA`, not `TRUE`, so that row gets dropped

```{r}
hrs_03_rm_poor |> tabyl(srh)
```
:::
:::

## Use `filter_out()` to keep `NA`s

::: columns
::: {.column width="60%"}

- We can use `filter_out()` from the `dplyr` package to keep the `NA`s when we want to perform an exclusionary (`!=`) filter

```{r}
hrs_04_rm_poor <- hrs_00 |>
  filter_out(srh == "Poor")

hrs_04_rm_poor |> tabyl(srh)
```

:::
::: {.column width="40%"}

 

::: pink
::: pink-ttl
Build this habit
:::
::: pink-cont
Whenever you filter on a column that has missing values, stop and check: *how many observations were removed and does that correspond to the number of values I expect (excluding `NA`s?)*
:::
:::
:::
:::

# Replace missing values with `replace_na()` {visibility="hidden"}

## `replace_na()`: replace missing values with a specific value

- Sometimes we want to fill in missing values with a specific value instead of removing the row
- We can use the `replace_na()` function *inside `mutate()`* to set `NA`s to a specific value
- `replace_na()` works on *numeric* and *character* variables, not *factors*
 

- Here's the general version of the function:

```{r}
#| eval: false
data_filled <- data |>
  mutate(
    variable_filled = replace_na( #<1>
      variable, #<2>
      value_to_fill_in #<3>
      )
  )
```

1. `replace_na()` swaps out any `NA` in `variable` for the value we provide. It's good practice to assign this to a new variable, like `variable_filled`
2. `variable` is the column we want to fill in
3. `value_to_fill_in` is the value we want to use to replace the missing

## Example 1: filling a numeric variable

- Suppose if `height` is missing, we want to replace it with the average `height` across the dataset
  - This is **not** an advised solution, I just want to show you how the function works!

- Here's how we can use `replace_na()` to fill in the missing values:

```{r}
hrs_05_height_filled <- hrs_00 |>
  mutate( #<1>
    height_filled = replace_na( #<2>
      height, #<3>
      mean(height, na.rm = TRUE)) #<4>
    ) #<1>
```

1. We need to wrap `replace_na()` in `mutate()` to create a new variable, `height_filled`
2. `replace_na()` swaps out any `NA` in `height` for the value we provide
3. `height` is the column we want to fill in.  
4. We will replace the missing values with the mean of `height`


```{r}
#| code-fold: true
#| code-summary: "Check that the function worked"

hrs_05_height_filled |>
  select(contains("height")) |>
  arrange(height) |> tail()
```

## `fct_na_value_to_level()`: filling a factor variable

- This is the `forcats` version of `replace_na()` for *factor* variables

 

- We can use the `fct_na_value_to_level()` function *inside `mutate()`* to set `NA`s to a specific value
- `fct_na_value_to_level()` works on *factors*!!

- General version of the function is similar to `replace_na()`:

```{r}
#| eval: false
data_filled <- data |>
  mutate( #<1>
    variable_filled = fct_na_value_to_level( #<2>
      variable, #<3>
      level = "new level to replace NAs" #<4>
      )
  ) #<1>
```

1. We need to wrap `fct_na_value_to_level()` in `mutate()` to create a new variable, `variable_filled`
2. `fct_na_value_to_level()` swaps out any `NA` in `variable` for the level we provide
3. `variable` is the column we want to fill in
4. `"new level to replace NAs"` is added as a new level, and used to replace the missing values

## Example 2: filling a factor variable

- Suppose if `race_original` is missing, we want to replace it with `Unknown`
  - This is **not** an advised solution, I just want to show you how the function works!

- Here's how we can use `replace_na()` to fill in the missing values:

```{r}
hrs_06_race_filled <- hrs_00 |>
  mutate(
    race_filled = fct_na_value_to_level( #<1>
      race_original, #<2>
      level = "Unknown") #<3>
    )
```

1. `fct_na_value_to_level()` swaps out any `NA` in `race_original` for the level we provide
2. `race_original` is the column we want to fill in
3. `"Unknown"` is added as a new level, and used to replace the missing values


```{r}
#| code-fold: true
#| code-summary: "Check that the function worked"

hrs_06_race_filled |> tabyl(race_original, race_filled)
```

# Replacing a factor level with `NA` {visibility="hidden"}

## `replace_with_na()`: replace a factor level with `NA`

- Sometimes there are data values that are not missing, but we want to treat them as missing
  - Example: you might see a numeric value as `-99` or `"999"` to indicate missingness, but because it's a valid number, R doesn't treat it as `NA`
- **Note:** there are no good examples of this in the HRS dataset, so I'm going to show you how to use the function with levels and values that **should not be considered missing**

 

- General version of the function:
```{r}
#| eval: false

data_replaced_na <- data |>
  replace_with_na(replace = list( #<1>
    numeric_var1 = numeric_val1, #<2>
    numeric_var2 = c(numeric_val2a, numeric_val2b, numeric_val2c), #<3>
    factor_var1 = c("factor_level1", "factor_level2")  #<4>
  ))
```

1. `replace_with_na()` takes a named list of columns and values to replace with `NA` 
2. `numeric_var1` is the name of a numeric column, and `numeric_val1` is the value to replace with `NA`
3. `numeric_var2` is the name of a numeric column, and `numeric_val2a`, `numeric_val2b`, and `numeric_val2c` are the values to replace with `NA`
4. `factor_var1` is the name of a factor column, and `"factor_level1"` and `"factor_level2"` are the levels to replace with `NA` 

## Example 3: replace certain levels or values with `NA` (1/2)

- We can replace a few values in the `hrs_06_race_filled` dataset with `NA`
```{r}
hrs_07_other_na <- hrs_06_race_filled |>
  mutate( #<1>
    sex_na = sex, 
    height_na = height,
    race_filled_na = race_filled
  ) |> #<1> 
  replace_with_na(replace = list( #<2>
    sex_na = "Male", #<3>
    height_na = 1.6002, #<4>
    race_filled_na = c("Other", "Unknown") #<5>
  ))
```

1. I often make a new variable to hold the original values, so I can compare before and after
2. `replace_with_na()` takes a named list of columns and values to replace with `NA`
3. In the `sex` column, we will replace `"Male"` with `NA`
4. In the `height` column, we will replace `70` with `NA`
5. In the `race_filled` column, we will replace `"Other"` and `"Unknown"` with `NA`


## Example 3: replace certain levels or values with `NA` (2/2)

- We need to check our data to make sure we replaced the values we wanted to replace with `NA`

::: columns
::: {.column width="28%"}
```{r}
#| code-fold: true
#| code-summary: "Check `sex_na`"

hrs_07_other_na |> 
  tabyl(sex, sex_na) |>
  gt() |>
  tab_options(table.font.size = px(35))
```
:::
::: {.column width="72%"}
```{r}
#| code-fold: true
#| code-summary: "Check `race_filled_na`"

hrs_07_other_na |> 
  tabyl(race_filled, race_filled_na) |>
  gt() |>
  tab_options(table.font.size = px(35))
```
:::
:::

```{r}
#| code-fold: true
#| code-summary: "Check `height_na`"

hrs_07_other_na |> 
  filter(is.na(height_na)) |>
  select(height, height_na) |>
  head()
```



# Wrap-up {visibility="hidden"}

## Wrap-up

- We learned a few ways to *handle* missing data, once we know it's there

 

- `drop_na()` removes rows with missing data — target specific columns, or drop any row with any missingness
- `filter()` **silently drops `NA` rows** unless you explicitly keep them with `| is.na(...)`; `filter_out()` handles this for you automatically
- `replace_na()` fills in missing values for **numeric** and **character** columns
- `fct_na_value_to_level()` fills in missing values for **factor** columns, by adding a new explicit level
- `replace_with_na()` goes the other direction: turns a specific *value* (or factor level) into `NA`
  - Always check your work afterward — compare before/after values with `tabyl()`, or compare `miss_var_summary()` before and after

 

- There is no single "correct" way to handle missing data — the right choice depends on **why** the data is missing and what you're trying to learn from it

## Resources

- [`drop_na()` reference](https://tidyr.tidyverse.org/reference/drop_na.html)
- [`filter_out()` reference](https://dplyr.tidyverse.org/reference/filter.html)
- [`replace_na()` reference](https://tidyr.tidyverse.org/reference/replace_na.html)
- [`fct_na_value_to_level()` reference](https://forcats.tidyverse.org/reference/fct_na_value_to_level.html)
- [`replace_with_na()` reference](http://naniar.njtierney.com/reference/replace_with_na.html)
- [Missing values chapter in *R for Data Science (2e)*](https://r4ds.hadley.nz/missing-values.html)
- [Dealing with missing data in the *Epidemiologist R Handbook*](https://epirhandbook.com/en/missing-data.html)
