---
title: "Data Summarization: Quick summaries"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Data Summarization: Quick summaries"
    html-math-method: mathjax
    highlight-style: ayu
    code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

pacman::p_load(
  here, 
  tidyverse, 
  writexl, 
  rstatix, 
  janitor, 
  naniar,
  gt,
  gtsummary,
  readxl, 
  rio, 
  skimr
)
```

```{r}
#| label: "import data" 
#| echo: false
hrs_data_rds <- import(here("data", "hrs_data.rds"))
hrs_00 <- clean_names(hrs_data_rds) |>
  relocate(diab, degree, .after = id)
```

## Quick summaries

- A good first step when working with data is to **explore the dataset, variables, and observations**
  - Does every variable have the expected number of observations?
  - Are there any missing values?
  - Are there any weird values that don't make sense?
- We want to **quickly browse the data** and get summary statistics for each variable

 

- There are a few functions in R that give you automatic summaries of your data:
  - `summary()` in base R
  - `skim()` in `skimr`
  - `get_summary_stats()` in `rstatix`

## What does `summary()` do?

- `summary()` will show:
  - For categorical data: the counts in each category
  - For numeric data: the minimum, 1st quartile, median, mean, 3rd quartile, maximum, and count for NA's

 

:::: pink
::: pink-cont
- No package needed: it's built into base R, so it always worksks
- Downside: no standard deviation, and the output is harder to read/scroll through for datasets with many columns
- Often what I use when I'm feeling super lazy, but **not the best for a first look at a whole dataset**
:::
::::

## What does `summary()` output look like?

**Input is a data frame, output is a list of summary statistics for each variable**

```{r}
summary(hrs_00)
```

**You can input a specific variable:**

```{r}
summary(hrs_00$age_yr) #<1>
```

1.  Use `$` to select a specific variable from a data frame

## What does `skim()` do?

- `skim()` will show a few tables:
  - Data summary: number of rows, number of columns, and types of variables
  - Variable type summaries:
    - For categorical data: the counts in each category, the number of unique values, and the most common value
    - For numeric data: the minimum, 1st quartile, median, mean, 3rd quartile, maximum, standard deviation, and count for NA's

 

:::: pink
::: pink-cont
- Best of the three for a first look at a whole dataset: nicely formatted, includes standard deviation, and separates variables by type automatically
- Downside: the output is a special `skim_df` object, so it's not as
:::
::::

## What does `skim()` output look like? {.scrollable}

```{r}
skim(hrs_00)
```

## What does `get_summary_stats()` do?

- `get_summary_stats()` will show a table of summary statistics for each numeric variable

 

:::: pink
::: pink-cont
- Only works for numeric variables: categorical variables are silently dropped
- Returns a data frame (a tibble), which makes it easy to use the results in further code, tables, or plots: this is its biggest advantage over `summary()` and `skim()`
:::
::::

## What does `get_summary_stats()` output look like?

```{r}
hrs_00 |>
  get_summary_stats()
```

## Flexibility of `get_summary_stats()`: show certain variables

- You can use the same syntax as the `select()` in `dplyr` to keep or remove specific columns
  - Use `variable_name` to keep and `-variable_name` to remove

```{r}
hrs_00 |>
  get_summary_stats(-birthmo)
```

## Flexibility of `get_summary_stats()`: show certain summary statistics

- Use `type =` to specify which summary statistics you want to see
  - Options include: `"common"`, `"full"`, `"mean_sd"`, `"mean_ci"`, `"median_iqr"`, and more (see `?get_summary_stats` for the full list)

```{r}
hrs_00 |>
  get_summary_stats(type = "common")
```

## Which one should I use?

+-----------------------+---------------------------------------------------------------------------+----------------------------------------------+
| Function              | Best for                                                                  | Watch out for                                |
+:======================+:==========================================================================+:=============================================+
| `summary()`           | -A fast, no-setup check of a small dataset                                | -No standard deviation                       |
|                       |                                                                           |                                              |
|                       |                                                                           | -Output can be hard to read for many columns |
+-----------------------+---------------------------------------------------------------------------+----------------------------------------------+
| `skim()`              | -A first, fairly thorough look at a whole dataset (categorical + numeric) | -Requires `skimr`                            |
|                       |                                                                           |                                              |
|                       |                                                                           | -Output isn't a plain data frame             |
+-----------------------+---------------------------------------------------------------------------+----------------------------------------------+
| `get_summary_stats()` | -Numeric summaries you want to reuse (e.g., in a table or plot)           | -Requires `rstatix`                          |
|                       |                                                                           |                                              |
|                       | -Grouped summaries (to be covered later)                                  | -Numeric variables only                      |
+-----------------------+---------------------------------------------------------------------------+----------------------------------------------+

 

- In practice, you'll often start with `skim()` to explore
- Then use `get_summary_stats()` if you have certain numeric variables and statistics you actually need

