---
title: "Data Summarization: Grouped summaries"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Data Summarization: Grouped summaries"
    html-math-method: mathjax
    highlight-style: ayu
    code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

pacman::p_load(
  here, 
  tidyverse, 
  writexl, 
  rstatix, 
  janitor, 
  naniar,
  gt,
  gtsummary,
  readxl, 
  rio, 
  skimr
)
```

```{r}
#| label: "import data" 
#| echo: false
hrs_data_rds <- import(here("data", "hrs_data.rds"))
hrs_00 <- clean_names(hrs_data_rds) |>
  relocate(diab, degree, .after = id)
```

## Grouped summaries

- Sometimes we want to compare summary statistics for different groups of people in our dataset
- You can use `group_by()` to specify the grouping variable(s) before summarizing
- Then you can use another summary function to calculate the statistics for each group separately

- We will review the functions that can be used for grouped summaries:
  - `summarise()` from `dplyr`
  - `skim()` from `skimr`
  - `get_summary_stats()` from `rstatix`
  - `tbl_summary()` from `gtsummary`
    - Slightly different: does not use `group_by()`
  
 

- For each, we will look at grouped summaries for:
  - Grouped by `sex`
  - Summarize `age_yr`, `cesd`, and `income`

## New summary function: `summarise()`/`summarize()`

- Not necessarily grouped
- Flexible function that allows you to calculate any summary statistic you want, and name the resulting columns yourself
  - You need to feed it a column and a function to apply to that column
  
 

::: pink
::: pink-cont
- Really flexible: you can calculate custom summary statistics
- More work if you want general summary statistics: you have to write out each statistic and name each resulting column yourself
:::
:::

## Example: summary statistics on age

- For each summary statistic, I need to use the **designated function** and **specify the column** I want to summarize

```{r}
hrs_00 |>
  summarise(
    mean_age = mean(age_yr, na.rm = TRUE),
    mean_age_above_80 = mean(age_yr[age_yr > 80], na.rm = TRUE), #<1>
    median_age = median(age_yr, na.rm = TRUE),
    sd_age = sd(age_yr, na.rm = TRUE),
    min_age = min(age_yr, na.rm = TRUE),
    max_age = max(age_yr, na.rm = TRUE), 
  )
```

1. I can create a unique summary statistic without changing the dataset. I have calculated the mean age for people above 80 years old.

## Group by a variable in `summarise()`

```{r}
hrs_00 |>
  group_by(sex) |> #<1>
  summarise( #<2>
    mean_age = mean(age_yr, na.rm = TRUE),
    sd_age = sd(age_yr, na.rm = TRUE),
    mean_cesd = mean(cesd, na.rm = TRUE),
    sd_cesd = sd(cesd, na.rm = TRUE),
    mean_income = mean(income, na.rm = TRUE),
    sd_income = sd(income, na.rm = TRUE)
  ) #<2>
```

1. Use `group_by()` to specify the grouping variable before summarizing. The summary statistics will be calculated separately for each level of `sex`.
2. Use `summarise()` to calculate the summary statistics. Syntax will be the same as non-grouped summaries.

## Group by a variable in `skim()` {.scrollable}

```{r}
hrs_00 |>
  group_by(sex) |> #<1>
  select(age_yr, cesd, income) |> #<2>
  skim() #<3>
```

1. Use `group_by()` to specify the grouping variable before summarizing. The summary statistics will be calculated separately for each level of `sex`.
2. Use `select()` to specify the variables you want to summarize.
3. Use `skim()` to calculate the summary statistics. Syntax will be the same as non-grouped summaries.

## Group by a variable in `get_summary_stats()`

```{r}
hrs_00 |>
  group_by(sex) |> #<1>
  get_summary_stats( #<2>
    age_yr, cesd, income, 
    type = "common"
    ) #<2>
```

1. Use `group_by()` to specify the grouping variable before summarizing. The summary statistics will be calculated separately for each level of `sex`.
2. Use `get_summary_stats()` to calculate the summary statistics. Syntax will be the same as non-grouped summaries. 

## Group by a variable in `tbl_summary()`

- Slightly different syntax than the other functions: does not use `group_by()`
- `by =` splits the table into side-by-side columns for each group

```{r}
hrs_00 |>
  select(age_yr, cesd, income, sex) |>
  tbl_summary(by = sex)                                    # <1>
```
1. Stratify the whole table by `sex`, creating one column per group

## Wrap-up

- We have covered several functions for summarizing **grouped data** in R
  - `summarise()`
  - `skim()`
  - `get_summary_stats()`
  - `tbl_summary()`

