---
title: "R Functions in Course"
description: "This resource contains all the R functions that we use in this class, including the lesson they are introduced and appear in."
date: today
format: 
  html:
    page-layout: full
    link-external-newwindow: true
    toc: true
    highlight-style: arrow
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

pacman::p_load(
  tidyverse,
  DT,
  here
)

functions_df <- read_csv(here("resources", "r_functions.csv"), show_col_types = FALSE) |>
  mutate(
    R_Documentation = if_else(
      !is.na(R_Documentation) & R_Documentation != "",
      paste0('<a href="', R_Documentation, '" target="_blank">docs</a>'),
      ""
    ),
    Additional_Help = if_else(
      !is.na(Additional_Help) & Additional_Help != "",
      paste0('<a href="', Additional_Help, '" target="_blank">link</a>'),
      "—"
    ),
    Function = paste0("<code>", Function, "()</code>")
  ) |>
  select(
    Function,
    Package,
    `Lesson #` = Lesson_Introduced,
    `Lesson Introduced` = Lesson_Topic,
    Documentation = R_Documentation,
    `Additional Help` = Additional_Help,
    `Appears in Lessons` = Appears_In_Lessons
  )
```

::: {.callout-note}
## Notes on this table

- Click a column header to **sort**, and use the search boxes under each header to **filter** by a column. The global search box (top right) searches across **all columns** at once.
- **Lesson #** / **Lesson Introduced** reflect the first lesson (per the course [schedule](https://nwakim.github.io/PUBH_523_26Su/schedule.html)) where the function is used in a code chunk or explicitly named in the slides.
- **Appears in Lessons** lists every lesson number where the function shows up again.
:::


```{r}
#| label: table

datatable(
  functions_df,
  escape = FALSE,               # allow the HTML links/code tags to render
  rownames = FALSE,
  filter = "top",                # per-column filter boxes
  extensions = "Buttons",
  options = list(
    pageLength = 25,
    lengthMenu = list(c(15, 25, 50, 100, -1), c("15", "25", "50", "100", "All")),
    dom = "Blfrtip",
    buttons = c("copy", "csv", "excel"),
    columnDefs = list(
      list(width = "110px", targets = 0),
      list(width = "90px",  targets = c(2, 4, 5))
    )
  ),
  class = "display compact stripe hover"
)
```
