---
title: "R Functions in Course"
description: "This resource contains all the R functions that we use in this class, including the lesson they are introduced and appear in."
date: today
format: 
  html:
    page-layout: full
    link-external-newwindow: true
    toc: true
    highlight-style: arrow
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

pacman::p_load(
  tidyverse,
  DT,
  here
)

# Base URL of the course site (used to build links to each lesson's slides)
site_url <- "https://nwakim.github.io/PUBH_523_26Su"

# Look up each lesson's slides from the lessons folder (e.g., "04_Basics_operators"),
# so links stay correct if lessons are renumbered again
lesson_links <- tibble(folder = basename(list.dirs(here("lessons"), recursive = FALSE))) |>
  filter(str_detect(folder, "^\\d+_")) |>
  mutate(
    Lesson_Introduced = as.numeric(str_extract(folder, "^\\d+")),
    Slides_URL = paste0(site_url, "/lessons/", folder, "/",
                        str_remove(folder, "^\\d+_"), ".html")
  ) |>
  select(Lesson_Introduced, Slides_URL)

functions_df <- read_csv(here("resources", "r_functions.csv"), show_col_types = FALSE) |>
  left_join(lesson_links, by = "Lesson_Introduced") |>
  mutate(
    R_Documentation = if_else(
      !is.na(R_Documentation) & R_Documentation != "",
      paste0('<a href="', R_Documentation, '" target="_blank">docs</a>'),
      ""
    ),
    Additional_Help = if_else(
      !is.na(Additional_Help) & Additional_Help != "",
      paste0('<a href="', Additional_Help, '" target="_blank">link</a>'),
      "—"
    ),
    Function = paste0("<code>", Function, "()</code>")
  ) |>
  select(
    Function,
    Package,
    `Lesson #` = Lesson_Introduced,
    `Lesson Introduced` = Lesson_Topic,
    Documentation = R_Documentation,
    `Additional Help` = Additional_Help,
    `Appears in Lessons` = Appears_In_Lessons,
    Slides_URL                     # hidden; used to link the lesson number
  )
```

::: {.callout-note}
## Notes on this table

- Click a column header to **sort**, and use the search boxes under each header to **filter** by a column. The global search box (top right) searches across **all columns** at once.
- **Lesson #** / **Lesson Introduced** reflect the first lesson (per the course [schedule](https://nwakim.github.io/PUBH_523_26Su/schedule.html)) where the function is used in a code chunk or explicitly named in the slides. Click the **Lesson #** to open that lesson's slides.
- **Appears in Lessons** lists every lesson number where the function shows up again.
:::


```{r}
#| label: table

datatable(
  functions_df,
  escape = FALSE,               # allow the HTML links/code tags to render
  rownames = FALSE,
  filter = "top",                # per-column filter boxes
  extensions = "Buttons",
  options = list(
    pageLength = 25,
    lengthMenu = list(c(15, 25, 50, 100, -1), c("15", "25", "50", "100", "All")),
    dom = "Blfrtip",
    buttons = c("copy", "csv", "excel"),
    columnDefs = list(
      list(width = "110px", targets = 0),
      list(width = "90px",  targets = c(2, 4, 5)),
      # Show the lesson number as a link to its slides, but keep it a number
      # underneath so sorting and the filter slider still work
      list(targets = 2, render = JS(
        "function(data, type, row) {",
        "  if (type === 'display' && row[7]) {",
        "    return '<a href=\"' + row[7] + '\" target=\"_blank\">' + data + '</a>';",
        "  }",
        "  return data;",
        "}"
      )),
      list(targets = 7, visible = FALSE, searchable = FALSE)
    )
  ),
  class = "display compact stripe hover"
)
```
