---
title: "Quarto and `.qmd`'s: Revisited"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Quarto and `.qmd`'s: Revisited"
    html-math-method: mathjax
    highlight-style: ayu
code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

pacman::p_load(
  tidyverse, 
  palmerpenguins,
  gt, 
  knitr,
  scales, 
  rio, 
  here, 
  janitor, 
  gtsummary
)

theme_set(theme_grey(base_size = 24)) 
```

```{r}
#| label: "import data" 
#| echo: false
hrs_data_rds <- import(here("data", "hrs_data.rds"))
hrs_00 <- clean_names(hrs_data_rds) |>
  relocate(diab, degree, .after = id)
```

## Recap and today's goals

- **Last time** (and what we've been using), we covered the building blocks of a `.qmd`:

  - YAML metadata, text, and code chunks
  - Creating, editing, saving, and rendering a file

 

- **Today**, we'll level up with tools that are especially useful when writing and styling reports:

  - Inline code
  - LaTeX math
  - Code chunks: cross-references and code folding
  - Fenced divs: callouts and tabsets

## Inline code: mixing R results into text

- Inline code is helpful for reports and updates: your write-up **automatically updates** if your data or code change!!
- Instead of typing a number by hand, you can pull it **directly from your code** using inline code

::: blue 
::: blue-cont 
[Written in `qmd`:]{style="color: #5BAFF8; font-weight: bold;"}

```` markdown
The HRS dataset has `r knitr::inline_expr("nrow(hrs_00)")` rows and `r knitr::inline_expr("n_distinct(hrs_00$hhid)")` 
households.
````
:::
:::

::: red
::: red-cont
[Which renders as:]{style="color: #C83532; font-weight: bold;"}

The HRS dataset has `r nrow(hrs_00)` rows and `r n_distinct(hrs_00$hhid)` households.
:::
:::

 

::: pink
::: pink-cont
Inline code uses single backticks with a lowercase `r` right after the first backtick (no curly braces, unlike a chunk)
:::
:::

## Inline code: helpful pairings

- `round()`:control decimal places
- `comma()`: format big numbers with commas (in `scales` package)
- `percent()`: format proportions as `%` (in `scales` package)

**Code it in R and assign to objects:**
```{r}
mean_income <- mean(hrs_00$income, na.rm = T) |> round(1) |> comma()
perc_sleep <- mean(hrs_00$sleep == "Yes", na.rm = T) |> percent(accuracy = 0.1)
```

::: blue
::: blue-cont
[Written in `qmd`:]{style="color: #5BAFF8; font-weight: bold;"}
```` markdown
The mean income in the HRS dataset is $`r knitr::inline_expr("mean_income")` and the percent 
of responders who have trouble sleeping is `r knitr::inline_expr("perc_sleep")`. 
````
:::
:::

::: red
::: red-cont
[Which renders as:]{style="color: #C83532; font-weight: bold;"}

The mean income in the HRS dataset is $`r mean_income` and the percent of responders who have trouble sleeping is `r perc_sleep`.
:::
:::


## LaTeX math: writing equations

Quarto renders **LaTeX math** using MathJax — no extra setup needed

- **Inline math**: wrap with single dollar signs — `$...$`
- **Display math**: wrap with double dollar signs — `$$...$$`

::: blue
::: blue-cont
[Written in `qmd`:]{style="color: #5BAFF8; font-weight: bold;"}

```` markdown
The slope $\beta_1$ describes the change in $Y$ for a one-unit change in $X$.

$$
Y_i = \beta_0 + \beta_1 X_i + \epsilon_i
$$
````
:::
:::

::: red
::: red-cont
[Which renders as:]{style="color: #C83532; font-weight: bold;"}

The slope $\beta_1$ describes the change in $Y$ for a one-unit change in $X$.

$$
Y_i = \beta_0 + \beta_1 X_i + \epsilon_i
$$
:::
:::

## LaTeX math: a worked example

We can use LaTeX math to write equations for the mean and proportion of a variable in our dataset

::: blue
::: blue-cont
[Written in `qmd`:]{style="color: #5BAFF8; font-weight: bold;"}

```` markdown
$$
\overline{\text{income}} = \frac{1}{n}\sum_{i=1}^{n} \text{income}_i
$$

$\overline{\text{income}} = \text{$} `r knitr::inline_expr("mean_income")`$
````
:::
:::

::: red
::: red-cont
[Which renders as:]{style="color: #C83532; font-weight: bold;"}

$$
\overline{\text{income}} = \frac{1}{n}\sum_{i=1}^{n} \text{income}_i
$$

$\overline{\text{income}} = \text{\$}`r mean_income`$
:::
:::

## Previously: code chunk options

You can control how a chunk behaves with `#|` options at the top of the chunk:

| Option | What it does |
|--------|--------------|
| `echo: false` | Hides the code in output (results still show) |
| `warning: false` | Hides warnings from the output |
| `include: false` | Hides both code and results (still runs) |
| `eval: false` | Don't run the code (no output) |

```{{r}}
#| echo: false #<1>
#| warning: false 
#| eval: false #<2>
```

1. This is helpful if you do not want to show your code
2. This is helpful if your `qmd` file has code that doesn't work or takes a long time to run, but you still want to render the document

## More code chunk options: code folding

- **Code folding** lets readers choose whether to see your code
- Keeps reports cleaner 

::: blue
::: blue-cont
[Written in `qmd`:]{style="color: #5BAFF8; font-weight: bold;"}

```` markdown
```{{r}}
#| code-fold: true
#| code-summary: "Calculating mean income and percent trouble sleeping"

mean_income <- mean(hrs_00$income, na.rm = T) |> round(1) |> comma()
perc_sleep <- mean(hrs_00$sleep == "Yes", na.rm = T) |> percent(accuracy = 0.1)
```
````
:::
:::

::: red
::: red-cont
[Which renders as:]{style="color: #C83532; font-weight: bold;"}
```{r}
#| code-fold: true
#| code-summary: "Calculating mean income and percent trouble sleeping"

mean_income <- mean(hrs_00$income, na.rm = T) |> round(1) |> comma()
perc_sleep <- mean(hrs_00$sleep == "Yes", na.rm = T) |> percent(accuracy = 0.1)
```
:::
:::

## More code chunk options: cross referencing

- You can **label** a code chunk with a figure or table, then reference it anywhere with `@`
- Labels must start with a **prefix**: `fig-`, `tbl-`, `eq-`, or `sec-`
- Quarto auto-numbers them and updates the numbers if you reorder slides

 

- We format our code chunk with a label and a figure caption (caption is "Plot of _____")
```` markdown
```{{r}}
#| label: fig-summ-stats
#| fig-cap: "Plot of _____"

# Code for plot goes here!
```

Then we reference inline with @fig-summ-stats
````

## More code chunk options: Example for cross referencing {.scrollable}

If we want to create a table of summary statistics, we can use the label to reference it later in the text:

::: {.panel-tabset}
### [Written in `qmd`:]{style="color: #5BAFF8; font-weight: bold;"}

```` markdown
```{{r}}
#| label: tbl-summ-stats
#| fig-cap: "Summary statistics for income and trouble sleeping"

hrs_00 |> 
  select(income, sleep) |>
  tbl_summary(
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  )
```

In @tbl-summ-stats, we see the mean and standard deviation for income and 
the proportion of respondents who report trouble sleeping.
````
 

### [Which renders as:]{style="color: #C83532; font-weight: bold;"}

```{r}
#| label: tbl-summ-stats
#| fig-cap: "Summary statistics for income and trouble sleeping"

hrs_00 |> 
  select(income, sleep) |>
  tbl_summary(statistic = list(all_continuous() ~ "{mean} ({sd})")) |>
  as_gt() |>
  tab_options(table.font.size = "40px")
```

In @tbl-summ-stats, we see the mean and standard deviation for income and the proportion of respondents who report trouble sleeping.
:::

## Callout blocks: highlight key info

- **Callouts** are a "fenced div" that draws the reader's eye to important notes, warnings, or tips
- There are five types of callouts: `note`, `tip`, `warning`, `important`, and `caution`
- You also have options for the appearance type: `default`, `minimal`, and `simple`

 

- We use `:::` and `{.callout-<type>}` to create a callout block, then close it with `:::` at the end

::: blue
::: blue-cont
[Written in `qmd`:]{style="color: #5BAFF8; font-weight: bold;"}

```` markdown
::: {.callout-note appearance="minimal"}
This is a neutral, informational note.
:::
````
:::
:::

::: red
::: red-cont
[Which renders as:]{style="color: #C83532; font-weight: bold;"}

::: {.callout-note appearance="minimal"}
This is a neutral, informational note.
:::
:::
:::

## Tabsets 

- **Tabsets** are a "fenced div" that organizes multiple views with clickable tabs
- We use `::: {.panel-tabset}` to create a tabset, then close it with `:::` at the end
  - Each tab is created with a header (e.g., `### Plot` or `### Table`) and the content for that tab goes below the header

::: columns
::: {.column width="37%"}
::: blue
::: blue-cont
[Written in `qmd`:]{style="color: #5BAFF8; font-weight: bold;"}

```` markdown
::: {.panel-tabset}
### Plot
Your ggplot code here

### Table
Your gt table here
:::
````
:::
:::
:::

::: {.column width="3%"}
:::

::: {.column width="60%"}

::: red
::: red-cont
[Which renders as:]{style="color: #C83532; font-weight: bold;"}

::: {.panel-tabset}
### Plot
Your ggplot code here

### Table
Your gt table here
:::
:::
:::
:::
:::

## Wrap-up

- We covered a few more tools for writing reports in Quarto!

 

- **Inline code** keeps numbers in your prose synced with your analysis
- **LaTeX math** lets you write clean equations, inline or on their own line
- **Cross-references and code folding** are code chunk options that help readers navigate your work
- **Callouts and tabsets** are fenced divs that help you write clear, well-organized reports

## Resources

- [Quarto: Cross-references](https://quarto.org/docs/authoring/cross-references.html)
- [Quarto: Callout blocks](https://quarto.org/docs/authoring/callouts.html)
- [Quarto: Tabsets](https://quarto.org/docs/presentations/revealjs/index.html#tabsets)
- [Quarto: The Practical Guide](https://quarto-tdg.org/) by Mine Çetinkaya-Rundel and Charlotte Wickham
  - The section on [Mastering Quarto: Markdown](https://quarto-tdg.org/authoring.html) is especially helpful for reprots