---
title: "Data visualization: Design choices"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Data visualization: Design choices"
    html-math-method: mathjax
    highlight-style: ayu
code-annotations: hover
execute:
  echo: true
  freeze: auto
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

pacman::p_load(
  here, 
  tidyverse, 
  writexl, 
  rstatix, 
  janitor, 
  naniar,
  gt,
  gtsummary,
  readxl, 
  rio, 
  skimr
)

theme_set(theme_grey(base_size = 24)) 
```

```{r}
#| label: "import data" 
#| echo: false
hrs_data <- import(here("data", "hrs_data.rds"))
hrs_00 <- clean_names(hrs_data) |>
  relocate(diab, srh, .after = id) |>
  drop_na() |>
  mutate(
    srh = fct_relevel(srh, "Poor", "Fair", "Good", "Very Good", "Excellent")
  )
```

## Recall: common components of `ggplot2`

::::: columns
::: column
- Common components of `ggplot2` include:
  - `ggplot()`: the main function to create a plot
  - `aes()`: the function to specify the mapping of variables to axes
  - `geom_*()` or `stat_*()`: functions to add layers to the plot
  - `labs()`: the function to change labels of axes and title
  - `theme()`: the function to adjust non-data plot elements such as text size,
:::

::: column
 

```{r}
#| eval: false

ggplot(
  data = my_data, 
  mapping = aes(x = col1, y = col2)
  ) +  #<1>       
  
  geom_*() +          
  
  labs() +       
  
  theme()                                  
```
:::
:::::

:::: blue
::: blue-cont
In this lesson, we will focus on **labels**, **themes**, and **facets**.
:::
::::

## Basic labelling using `labs()`: titles and axes

- We can assign many different labels to a plot using the `labs()` function

We have the following options for labelling:

```{r}
#| eval: false

ggplot() +     
  geom_*() +          
  labs(
    x = "Label for x-axis", #<1>
    y = "Label for y-axis", #<1>
    color = "Label for color legend", #<2>
    fill = "Label for fill legend", #<2>
    title = "Title for the whole plot", #<1>
    subtitle = "Appears smaller, under the title", 
    caption = "Appears at the bottom of the plot, good for context" 
  ) +       
  theme()                                  
```

1.  Typically all you need for a plot with two variables.
2.  Only needed if you have a legend for color or fill.

## Example 1: New labels for our plot of age vs. self-reported health

- The following example adds all possible labels to the plot
- We do NOT need to add every possible label!!

```{r}
#| fig-height: 9
#| fig-width: 10
#| output-location: column

ggplot(
  hrs_00, 
  aes(x = age_yr, y = srh) #<3>
  ) +
  geom_boxplot() + #<4>
  geom_jitter(alpha = 0.2) + #<5>
  labs(
    x = "Age (years)", #<1>
    y = "Self-reported health", #<1>
    title = "Self-reported health by age", #<1>
    subtitle = "Data from the Health and Retirement Study (HRS)", #<1>
    caption = "Source: HRS, 2022 wave"
  )
```

## Example 2: Labels for color

- Note that the HRS dataset came with really nice labels embedded in the `.rda` file

::::: columns
::: column
```{r}
#| fig-height: 8
#| fig-width: 10
#| code-fold: true
#| code-summary: "Original labels"


ggplot(
  hrs_00,
  aes(x = age_yr, y = srh, color = srh)
  ) +
  geom_jitter()
```
:::

::: column
```{r}
#| fig-height: 8
#| fig-width: 10
#| code-fold: true
#| code-summary: "Some label changes"


ggplot(
  hrs_00,
  aes(x = age_yr, y = srh, color = srh)
  ) +
  geom_jitter() +
  labs(
    x = "Age (years)", #<1>
    color = "Self-reported \nhealth" #<2>
  )
```

1.  Take away "at Interview" for x-axis
2.  Added `\n` to condense the horizontal space that the legend title takes up.
:::
:::::

## Example 3: Same plot without our nice, pre-labelled variables

- If we use the `.xlsx` file, we lose the labels and the SRH level order.

```{r}
hrs_xlsx <- import(here("data", "hrs_data.xlsx")) |> drop_na()
```

::::: columns
::: column
```{r}
#| fig-height: 8
#| fig-width: 10
#| code-fold: true
#| code-summary: "Variable names are the labels"


ggplot(
  hrs_xlsx,
  aes(x = age_yr, y = srh, color = srh)
  ) +
  geom_jitter()
```
:::

::: column
```{r}
#| fig-height: 8
#| fig-width: 10
#| code-fold: true
#| code-summary: "Some label changes"


ggplot(
  hrs_xlsx,
  aes(x = age_yr, y = srh, color = srh)
  ) +
  geom_jitter() +
  labs(
    x = "Age (years)",
    y = "Self-reported health", 
    color = "Self-reported \nhealth" #<2>
  )
```

1.  Take away "at Interview" for x-axis
2.  Added `\n` to condense the horizontal space that the legend title takes up.
:::
:::::

## Facets

- Facets allow us to create multiple plots based on one or more discrete variables
  - Discrete meaning categorical or whole number values

 
  
- We create multiple plots based on subsets of the data
  - If each subset is determined by one variable, we use `facet_wrap()`
  - If each subset is determined by two variables, we use `facet_grid()`

## Example 4: `facet_wrap()`

- We use `srh` as the facet variable to create a histogram of age for each level of self-reported health

```{r}
#| fig-height: 9
#| fig-width: 10
#| output-location: column

ggplot(
  hrs_00,
  aes(x = age_yr)
  ) +
  geom_histogram() +
  facet_wrap(~ srh) #<1>
```

1.  The `~` indicates that we are faceting by a single variable, `srh`.

## Example 5: `facet_grid()`

- We use `srh` and `diab` as the facet variables to create a histogram of age for each level of self-reported health and diabetes status

```{r}
#| fig-height: 7
#| fig-width: 10
#| output-location: column

ggplot(
  hrs_00,
  aes(x = age_yr)
  ) +
  geom_histogram() +
  facet_grid(diab ~ srh) #<1>
```

1.  The `~` indicates that we are faceting by two variables, `diab` and `srh`. The first variable is used for the rows and the second variable is used for the columns.

## Complete themes

- We can change the entire appearance of our plot with **themes**
- There are several set themes in `ggplot2`: [check out the page on complete themes for more!](https://ggplot2.tidyverse.org/reference/ggtheme.html)

 

::::::: columns
::: {.column width="25%"}
```{r}
#| fig-height: 3
#| fig-width: 4
#| code-line-numbers: "3"

ggplot(hrs_00) +
  geom_histogram(aes(x = age_yr)) + 
  theme_grey() #<1>
```

1.  This is the default theme!
:::

::: {.column width="25%"}
```{r}
#| fig-height: 3
#| fig-width: 4
#| code-line-numbers: "3"

ggplot(hrs_00) +
  geom_histogram(aes(x = age_yr)) +
  theme_bw()
```
:::

::: {.column width="25%"}
```{r}
#| fig-height: 3
#| fig-width: 4
#| code-line-numbers: "3"

ggplot(hrs_00) +
  geom_histogram(aes(x = age_yr)) +
  theme_classic()
```
:::

::: {.column width="25%"}
```{r}
#| fig-height: 3
#| fig-width: 4
#| code-line-numbers: "3"

ggplot(hrs_00) +
  geom_histogram(aes(x = age_yr)) +
  theme_minimal()
```
:::
:::::::

## Modifying themes

- We can change various components of a theme using the `theme()` function

 

- We can change the look of the title, axes, facets, and legend
  - [The input options for `theme()` is VAST](https://ggplot2.tidyverse.org/reference/theme.html)

 

- We're only going to touch the surface here!

## Example 6: Modifying the title and axes {.scrollable}

```{r}
#| fig-height: 8
#| fig-width: 10
#| code-fold: true
#| code-summary: "Adding different theme elements"

ggplot(
  hrs_00, 
  aes(x = age_yr, y = srh) 
  ) +
  geom_boxplot() + 
  geom_jitter(alpha = 0.2) + 
  labs(
    x = "Age (years)", 
    y = "Self-reported health",
    title = "Self-reported health by age",
  ) +
  theme(
    plot.title = element_text(size = 30, face = "bold", color = "blue"), #<1>
    axis.title.x = element_text(size = 20, face = "italic", color = "red"), #<2>
    axis.title.y = element_text(size = 20, color = "red"),  #<3>
    axis.text.y = element_text(size = 16, angle = 45) #<4>
  )
```

1.  Changes the title size, makes it bold, and changes the color to blue
2.  Changes the x-axis title size, makes it italic, and changes the color to red
3.  Changes the y-axis title size and color to red
4.  Changes the y-axis text size and rotates it 45 degrees

## Example 7: Modifying the facets {.scrollable}

```{r}
#| fig-height: 7
#| fig-width: 10
#| code-fold: true
#| code-summary: "Adding different theme elements to facet"

ggplot(
  hrs_00,
  aes(x = age_yr)
  ) +
  geom_histogram() +
  facet_grid(diab ~ srh) +
  theme(
    strip.text.x = element_text(size = 16, color = "blue"), #<1>
    strip.text.y = element_text(size = 16, color = "red") #<2>
  )
```

1.  Changes the facet column label size and color to blue
2.  Changes the facet row label size and color to red

## Example 8: Modifying the legend {.scrollable}

```{r}
#| fig-height: 8
#| fig-width: 10
#| code-fold: true
#| code-summary: "Adding different theme elements to the legend"


ggplot(
  hrs_00,
  aes(x = age_yr, y = srh, color = srh)
  ) +
  geom_jitter() +
  labs(
    x = "Age (years)", 
    color = "Self-reported health"
  ) +
  theme(
    legend.position = "bottom" #<1>
  )
```

1.  Moving the legend to the bottom of the plot instead of the right (default)

## Saving plots

- We can save a plot as an image for later use: *way better quality than taking a screenshot of the plot*
- We need to assign the plot to an R object, and then we can save the R object as an image or pdf
- We use `ggsave()` to save an R object (that's a plot)

Let's say I saved my plot from Example 8 as `plot_age_srh`:

::::: columns
::: column
```{r}
plot_age_srh <- ggplot( #<1>
  hrs_00,
  aes(x = age_yr, y = srh, color = srh)
  ) +
  geom_jitter() +
  labs(
    x = "Age (years)", 
    color = "Self-reported health"
  ) +
  theme(
    legend.position = "bottom"
  )
```

1.  I assign the output (aka my plot) to `plot_age_srh`
:::

::: column
```{r}
ggsave(
  filename = here( #<1>
    "images", 
    "plot_age_srh.png"
    ), #<1>
  plot = plot_age_srh, #<2>
  width = 5, #<3>
  height = 3, #<3>
  units = "in",
  dpi = 200)
```

1.  We need to tell `R` where we want the image to end up in our files
2.  We give `R` the name of the plot
3.  We need to adjust the height and width to make the plot readable in the image
:::
:::::

## Wrap-up

- We went over additional layers for `ggplot2`

 

- We can use **labels** to make plots more readable to viewers
- We can use **facets** to wrap plots by subsets of data
- We can use **themes** to change the overall appearance of a plot
- We can use **`ggsave()`** to save plots as images

 

- Check out more on `ggplot2`: it has a lot of capabilities!

## Resources

- [`ggplot2` Cheatsheet in html form](https://rstudio.github.io/cheatsheets/html/data-visualization.html)
- [`ggplot2` Cheatsheet in pdf form](https://rstudio.github.io/cheatsheets/data-visualization.pdf)
- [List of complete themes](https://ggplot2.tidyverse.org/reference/ggtheme.html)
- [Modify components of a theme](https://ggplot2.tidyverse.org/reference/theme.html)
- [Page on `ggsave()`](https://ggplot2.tidyverse.org/reference/ggsave.html)
- Online textbook for `ggplot2`: <https://ggplot2-book.org/>
- Another online resource for data visualization with `ggplot2`: <https://socviz.co/index.html#preface>
- [Advanced Data Viz Class at Duke](https://vizdata.org/)

 

- If you found these lessons on Data Visualization fun, I highly suggest you look into Dr. Bedrick's [Data Visualization course](https://dmice.ohsu.edu/bedricks/courses/bmi525_spring_2025/) in the Spring!
