---
title: "Lab 1"
subtitle: "PUBH 523/623"
date-modified: "today"
format: 
  html:
    link-external-newwindow: true
    toc: true
    self-contained: true
    embed-resources: true
    number-sections: true
    number-depth: 4
    highlight-style: ayu
editor_options: 
  chunk_output_type: console
---

```{r}
#| message: false

# PLEASE DO NOT REMOVE THIS CODE CHUNK!!!
### ADD YOUR LIBRARIES HERE!!! ####
pacman::p_load( 
  here, 
  tidyverse, 
  writexl, 
  rstatix, 
  janitor, 
  naniar,
  gt,
  gtsummary,
  readxl
)
```

```{=html}
<style>
/* Hide numbers for H1 and H2 in the main body */
h1 .header-section-number, 
h2 .header-section-number {
    display: none;
}
</style>
```

## Directions {.unnumbered}

Please turn in your `.qmd` and `.html` file on Brightspace. Please use the same naming convention as practice assignments: `lastname_firstinit_lab_01.qmd` and `lastname_firstinit_lab_01.html`.

**Please do not remove anything from this file!!** You will only add your code and work to this file.

[The instructions for this lab can be found here.](https://nwakim.github.io/PUBH_523_26Su/project/Lab_01_instructions.html)

## Lab activities

### Familiarize yourself with the Well-Being and Basic Needs Survey

::: callout-important
#### Task

Answer the following questions using information on WBNS:

- What is the motivation for this study?
- How could an information sheet on components of the WBNS help change policy?
- Why is it important to study basic needs?
:::

**What is the motivation for this study?**

Type your answer here.

**How could an information sheet on components of the WBNS help change policy?**

Type your answer here.

**Why is it important to study basic needs?**

Type your answer here.

### File organization

::: callout-important
#### Tasks

- Display your working directory using the `here` package and `here()` function.

- Insert an image of your folder organization with the `.Rproj` file in the correct place. 
:::

```{r}
# Code to display your working directory here

```

Insert the image of your folder organization here.

### Access and download the data

::: callout-important
#### Task

No task to report back on. Just make sure you have the data and read the User Guide!
:::

### Load the data and rename the dataset

::: callout-important
#### Task

Load your downloaded data file into R. Rename the loaded object to a clean dataset name of your choice. (Make sure to show the code for this!)
:::

```{r}
# Code to load and rename your dataset here

```

### Decide on your project question

::: callout-important
#### Task

Write down your chosen project option number and question.
:::

**Project option number:** 

**Project question:** 

### Getting data in working format

::: callout-important
#### Task

Remove the parentheses with values that are in front of the category names. Verify that the parentheses are gone using `glimpse()`.
:::


```{r}
# Code to remove parentheses and then
# verify the parentheses are gone using glimpse() here

```

### Save the data 

::: callout-important
#### Task

Include a screenshot showing the new `.rds` file within your data folder.
:::

```{r}
# Code to save dataset here

```

Insert the screenshot of your `.rds` file in your data folder here.
