---
title: "Data Summarization: Tables"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Data Summarization: Tables"
    html-math-method: mathjax
    highlight-style: ayu
    code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

pacman::p_load(
  here, 
  tidyverse, 
  writexl, 
  rstatix, 
  janitor, 
  naniar,
  gt,
  gtsummary,
  readxl, 
  rio, 
  skimr
)
```

```{r}
#| label: "import data" 
#| echo: false
hrs_data_rds <- import(here("data", "hrs_data.rds"))
hrs_00 <- clean_names(hrs_data_rds) |>
  relocate(diab, degree, .after = id)
```

## What are summary tables?

- Summary tables are a way to present data in a structured format
- The use is a little vague here... 
  - It could mean:
    - A table for categorical data (like a frequency table)
    - A presentable table for categorical and numeric data

 

- We will cover two functions that can help us create summary tables for categorical and numeric data:
  - `tabyl()` from the `janitor` package
  - `tbl_summary()` from the `gtsummary` package

## What does `tabyl()` do?

- `tabyl()` is a simple function that creates a frequency table for a single categorical variable
- It returns a data frame with three columns:
  - `n`: the count in each category
  - `percent`: that count as a proportion of *all* rows (including missing values)
  - `valid_percent`: that count as a proportion of only the *non-missing* rows

::: pink
::: pink-cont
- Pretty simple to use and can be used to create two-by-two tables
- Need to feed it one variable at a time
:::
:::

## What does `tabyl()` output look like?

```{r}
hrs_00 |>
  tabyl(degree)
```

## Cross-tabulations with `tabyl()`

- You can also cross-tabulate two variables by adding a second column name
  - This will give you a table of counts for each combination of the two variables

```{r}
hrs_00 |>
  tabyl(degree, sex)
```

## Dressing up a `tabyl()` with `adorn_*()`

- `adorn_*()` functions let you format that table for presentation
- These are designed to be piped together, one after another

| Function | What it does |
|----|--------|
| `adorn_totals()` | Adds a total row and/or column (`where = "row"`, `"col"`, or `"both"`) |
| `adorn_percentages()` | Converts counts to proportions (`denominator = "row"`, `"col"`, or `"all"`) |
| `adorn_pct_formatting()` | Converts proportions to nicely formatted percents (control decimals with `digits =`) |
| `adorn_ns()` | Adds the raw counts back onto a table of percents, e.g. `"45.2% (120)"` |

## Example: Formatting a frequency table

- We can add totals, convert to percentages, format as percents, and add counts back to a frequency table
- Order matters! 
  - You generally want: tabulate → add totals → convert to percentages → format as percents → add counts back
  
```{r}
hrs_00 |>
  tabyl(degree) |>
  adorn_totals(where = "row") |> #<1>
  adorn_pct_formatting(digits = 2) #<2>
```

1. Add a row for totals to the table. A little confusing because it totals each column, but "row" refers to the position of the total row, not what is being totaled
2. Format the percentages to 2 decimal places

## Example: Formatting a cross-tabulation

- We can also add totals, convert to percentages, format as percents, and add counts back to a cross-tabulation

```{r}
hrs_00 |>
  tabyl(degree, sex) |> #<1>
  adorn_totals(where = "row") |> #<2>
  adorn_percentages(denominator = "col") |> #<3>
  adorn_pct_formatting(digits = 1) |> #<4>
  adorn_ns(position = "front") #<5>
```

1. Create a cross-tabulation of `degree` by `sex`
2. Add a row for totals to the table. A little confusing because it totals each column.
3. Convert the counts to percentages, using the column totals (the added row of totals) as the denominator.
4. Format the percentages to 1 decimal place
5. Add the raw counts back to the table, in front of the percentages.

## What does `tbl_summary()` do?

- `tbl_summary()` builds a publication-ready summary table for *all* the columns you give it in a single command
- It automatically picks sensible statistics based on each variable's type:
  - Numeric variables: median (interquartile range)
  - Categorical variables: count (percent)
  - Missing values are shown as their own "Unknown" row
- We need to use `select()` to decide what goes into the table

 

::: pink
::: pink-cont
- Great for publications!
- You can customize the statistics shown, the labels, and even stratify the table by a grouping variable
- Some of the customizations are syntax heavy
:::
:::

## What does `tbl_summary()` output look like? {.scrollable}

::: columns
::: column
```{r}
#| eval: false

hrs_00 |>
  select(-id, -hhid, -pn) |> #<1>
  tbl_summary()
```

1. Only removing this because it would add thousands of rows to this table! See what happens when you don't remove these variables!
:::
::: column
```{r}
#| echo: false

hrs_00 |>
  select(-id, -hhid, -pn) |>
  tbl_summary()
```
:::
:::

## Customizing `tbl_summary()`

- `statistic =` lets you change which statistics are shown, using `{ }` to insert stat names like `{mean}`, `{sd}`, `{median}`, `{n}`, `{p}` (percent)
- `label =` lets you rename variables to something more readable

```{r}
hrs_00 |>
  select(age_yr, sex, diab) |>
  tbl_summary(
    statistic = age_yr ~ "{mean} ({sd})",          # <1>
    label = age_yr ~ "Age (years)"                 # <2>
  )
```
1. Show mean (sd) for `age_yr` instead of the default median (IQR)
2. Give `age_yr` a nicer display name

## Presentation-ready tables with `gt()`

- `gt()` turns any data frame or table (including a `tabyl()` output) into a polished, presentation-ready table
- It's a general-purpose "make this table look nice" tool
- You can further customize titles, column labels, and colors using additional `gt` functions (like `tab_header()` or `fmt_percent()`)

::: columns
::: column
```{r}
#| code-fold: true
#| code-summary: "Original table from `tabyl()`"

hrs_00 |>
  tabyl(degree)
```
:::
::: column
```{r}
#| code-fold: true
#| code-summary: "Using `gt()` to make it look nice"

hrs_00 |>
  tabyl(degree) |>
  gt() |>
  tab_header(title = "Highest Degree Earned")
```
:::
:::

## We can change the font size for `tbl_summary()`

```{r}
hrs_00 |>
  select(age_yr, sex, diab) |>
  tbl_summary(
    statistic = age_yr ~ "{mean} ({sd})",          
    label = age_yr ~ "Age (years)"                 
  ) |>
  as_gt() |> #<1>
  tab_options(table.font.size = px(50)) #<2>
```

1. Convert the `tbl_summary()` output to a `gt` table so we can customize it. Need to use `as_gt()` instead of `gt()` because `tbl_summary()` already creates a table.
2. Change the font size to 50 pixels. Be careful: font size is relative to your document. 50px is probably too large for your standard html's. 

## Wrap-up

- We covered a few functions for summarizing data in R:
  - `tabyl()` from the `janitor` package
    - For categorical data
    - Can be used for one-way or two-way tables
  - `tbl_summary()` from the `gtsummary` package
    - For categorical and numeric data
    - Good for publications!
    - Can be customized to show different statistics, labels, and stratifications

## Resources

- [Tutorial on `tbl_summary()` from the `gtsummary` package](https://www.danieldsjoberg.com/gtsummary/articles/tbl_summary.html)
  - Includes more options than covered here!
- [More on the `gt` package](https://gt.rstudio.com/articles/gt.html)
  - Includes more options than covered here!
