---
title: "Data transformation: Rows"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Data transformation: Rows"
    html-math-method: mathjax
    highlight-style: ayu
    code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message_yr: false
#| warning: false

pacman::p_load(
  here, 
  tidyverse, 
  writexl, 
  rstatix, 
  janitor, 
  naniar,
  gt,
  gtsummary,
  readxl, 
  rio
)
```

```{r}
#| echo: false
hrs_data_rds <- import(here("data", "hrs_data.rds"))
hrs_data <- clean_names(hrs_data_rds) |>
  relocate(age_yr, ed, cesd, .after = birthdate) |>
  select(id, age_yr, ed, cesd, degree, srh)
```

## Data transformations on rows

- We defined `dplyr` functions as operations that use columns as arguments
  - This is still true for row transformations
  - We use columns to define the rows we want to keep, remove, or modify

 

- There are many functions that can be used to transform rows

 

::: columns
::: {.column width="50%"}
- We will cover the most common ones here: 
  - `filter()`
  - `arrange()`
  - `distinct()`
:::

::: {.column width="10%"}
:::

::: {.column width="40%"}
::: pink
::: pink-cont
Please note that I reduced the columns in `hrs_data` to make the output easier to read. The functions work the same way with all columns.
:::
:::
:::
::: 
  
# `filter()`
  
## `filter()`

- `filter()` allows us to keep rows (aka subset our data) based on conditions we specify (in the columns)

- We input our **data frame as the first argument**, and then we **specify the conditions for filtering in the second argument**

::: columns
::: {.column width="40%"}
- We can use the following comparison operators to specify our conditions:
    -   `>`: greater than
    -   `<`: less than
    -   `>=`: greater than or equal to
    -   `<=`: less than or equal to
    -   `==`: equal to
    -   `!=`: not equal to
:::
::: {.column width="60%"}
![[Artwork by \@allison_horst](https://allisonhorst.com/everything-else){target="_blank"}](../img_slides/horst_dplyr_filter.jpg){fig-align="center"}
:::
:::

## Example: filtering `age_yr` 

- Let's filter on a numeric variable, `age_yr`: all rows with `age_yr` values greater than 99
  - We will expect to return rows with `age_yr` values of 100 and above from `hrs_data`

```{r}
hrs_data |> # <1>
  filter( # <2>
    age_yr > 99 # <3>
    ) # <2>
```

1. The first argument is the data frame we want to operate on (in this case, `hrs_data`)
2. The second argument is the condition we want to apply to filter our data (in this case, `age_yr > 99`)
3. The condition is specified using a comparison operator (in this case, `>`) and the variable we want to filter on (in this case, `age_yr`)


## Example: filtering `age_yr`, now only exact matches

- Let's filter on a numeric variable, `age_yr`: all rows with `age_yr` values equal to 38
  - We will expect to return rows with `age_yr` values of 38 from `hrs_data`

```{r}
hrs_data |> 
  filter(age_yr == 38) # <1>
```

1. **Notice the change in the operator!!** We are now using `==`. 

## Example: filtering degree

- `degree` is a categorical variable
- For categorical variables, we can only use `==` or `!=` to filter on exact matches
- `degree` has the following possible values: `NA`, `None`, `GED`, `High school diploma`, `Associate's degree`, `Bachelor's degree`, and `Master's degree`, `Professional degree (Ph.D./M.D./JD)`

```{r}
#| output: false

hrs_data |>
  filter(degree == "Master's degree") # <1> 
```

1. We must include " " around the value for categorical variables. The exact spelling needs to match. 


```{r}
#| echo: false
hrs_data |>
  filter(degree == "Master's degree") |> 
  head()
```

## Example: filtering degree but mispelling the value

- What happens if I mispell "Master's degree" as "Masters degree" (without the apostrophe)?
- **R will not throw an error: we need to check our output!!**

```{r}
hrs_data |>
  filter(
    degree == "Masters degree" # <1> 
    )
```

1. I forgot the apostrophe in "Master's degree" and I get 0 rows returned. 

## Example: filtering degree for everything except a Master's

-   If we want everything* except Master's degrees, we use the "not equal" sign (`!=`)
    - *Everything does not include `NA` values, which we will discuss in the next slide

```{r}
hrs_data |>
  filter(degree != "Master's degree")
```

## `filter_out()`: new companion to `filter()`

-   `filter_out()` requires `{dplyr}` 1.2.0 (or later version)
- Allows us to filter **out** rows based on conditions we specify (in the columns)

```{r}
hrs_data |>
  filter_out(degree == "Master's degree") #<1>
```

1. We filter out anyone whose degree is a Master's degree. This is the same as using `filter(degree != "Master's degree")` but it keeps the `NA` values in the output.

## WAIT! Did those two operations have different rows?

- It's always good to check your output!

 

- I did not notice that `filter_out()` and `filter()` with `!=` had different outputs until I checked the number of rows in each output

 

- `filter_out()` keeps the `NA` values, while `filter()` with `!=` does not keep the `NA` values (because `NA` is not equal to anything, including itself)

::: columns
::: {.column width="25%"}
:::
::: {.column width="50%"}
::: blue
::: blue-cont
When using `!=`, I suggest using `filter_out()`
:::
:::
:::
::: {.column width="25%"}
:::
:::

## Filtering out missing values using `drop_na()`

-   We may want to remove rows that have missing data, which are coded as `NA`.

- If we want to remove rows with missing data in **ANY** of the columns
  - AKA, we are only keeping complete observations

```{r}
hrs_data |> drop_na()
```

## Filtering out missing values in specific columns using `drop_na()`

```{r}
hrs_data |> drop_na(degree) #<1>
```

1. This will only drop rows with `NA` values in the `degree` column, and will keep rows with `NA` values in other columns.

# `arrange()`

## **Sorting** datasets using `arrange()`

- `arrange()` changes the order of the rows in a dataset based on the values in one or more columns

 

- You can arrange by ascending or descending order

 

- Why use `arrange()`? 
  - Visual clarity when you look at data (see a variable ordered)
  - Helpful in conjunction with other functions (like `count()`)

## Example: sort data by ascending birth year

-   By default, `arrage_yr()` sorts in *ascending* order:

```{r}
hrs_data |> 
  arrange(age_yr)
```

## Example: sort data by descending birth year

```{r}
hrs_data |> 
  arrange(desc(age_yr)) #<1> 
```

1. We wrap `birthyr` in `desc()` to sort by descending birth year.

## Sorting datasets by multiple variables

-   You can also arrange by multiple variables.
    -   The code below sorts by `age_yr` *first*, and then within `age_yr`, we sort by `ed`.

```{r}
hrs_data |> 
  arrange(age_yr, ed) #<1>
```

1. This will sort by `age_yr` first, and then within `age_yr`, it will sort by `ed`.

## Sorting datasets by multiple variables: order matters!

```{r}
hrs_data |> 
  arrange(ed, age_yr) #<1>
```

1. This will sort by `ed` first, and then within `ed`, it will sort by `age_yr`. This is different from the previous code chunk where we sorted by `age_yr` first, and then within `age_yr`, we sorted by `ed`.

# `distinct()`

## `distinct()` keeps unique rows

- `distinct()` allows us to keep only unique rows based on the values in one or more columns

 

- `distinct()` can run without any arguments: it will return all unique rows in the dataset
  - This can be helpful if you want to perform a quality check on your dataset to see if there are any duplicate rows
  
 

- `distinct()` can also take specific columns as arguments: it will return unique rows based on the values in those columns

## Example: keep unique rows in dataset

```{r}
hrs_data |> 
  distinct()
```

## Example: keep unique rows based on `degree`

```{r}
hrs_data |> 
  distinct(degree) #<1>
```

1. This will return unique rows based on the values in the `degree` column. The output will only include the `degree` column, and it will return one row for each unique value in the `degree` column.

## Example: keep unique rows based on `degree` *with all columns*

```{r}
hrs_data |> 
  distinct(degree, .keep_all = TRUE) #<1>
```

1. If we want to keep all columns in the output, we can use the argument `.keep_all = TRUE`. This will return all columns for the unique rows based on the specified column(s).

## Example: keep unique rows based on `degree` and `self_health`

```{r}
hrs_data |> 
  distinct(degree, srh) #<1>
```

1. This will return unique rows based on the values in the `degree` and `self_health` columns. This allows us to see unique combinations of `degree` and `self_health` in our dataset.

## Wrap-up

- We learned three functions that allow us to transform rows in our dataset:
  - `filter()` (and `filter_out()`)
    - **Keep rows based on conditions** we specify in the columns
  - `arrange()`
    - **Sort rows based on the values** in one or more columns
  - `distinct()`
    - **Keep only unique rows** based on the values in one or more columns
