---
title: "Missing data: summarization and visualization"
author: "Nicky Wakim, Jessica Minnier, and Meike Niederhausen"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Missing data: summarization and visualization"
    html-math-method: mathjax
    highlight-style: ayu
    code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: setup
#| include: false

pacman::p_load(
  tidyverse, 
  here, 
  rio, 
  naniar, 
  gt
)

hrs_data <- import(here("data", "hrs_data.rds")) |>
  mutate(srh = fct_relevel(srh, "Excellent", "Very Good", "Good", "Fair", "Poor"))

theme_set(theme_grey(base_size = 50)) 
```

# Missing data summary and visualization {visibility="hidden"}

## Before we start

- Install and load the `naniar` package (a play on Narnia) if you haven't already!

![[Artwork by @allison_horst](https://allisonhorst.com/r-packages-functions)](../img_slides/horst_naniar.png){fig-alt="\"Explore missing data in naniar!\" Cute monsters with exploring gear looking around a landscape full of subtle (and not so subtle) NAs, hidden in trees, on mountain tops, as birds, and in flowers. Learn more about naniar."}


- We'll be working with only a few variables in the HRS dataset in this lesson!
  - Mostly to help us view the tables and plots within the slides
  
```{r}
hrs_00 <- hrs_data |>
  select(ed, race_original, srh, sex, act_vig, height)
```

 

## Why summarize and visualize missingness before doing anything else?

- Before you transform, model, or plot a variable, you need to know **how much of it is actually there**

 

::: columns
::: column
::: blue
::: blue-ttl
Missing data can be a problem for many reasons
:::
::: blue-cont
- It can *reduce your sample size* (and therefor your ability to make conclusions)
- It can *bias your results* if there's a pattern to missingness
- It can *make your plots look weird* or misleading
:::
:::
:::

::: column
::: pink
::: pink-ttl
A quick missingness check can reveal
:::
::: pink-cont
- Which *variables have the most missing data*
- Whether *missingness is scattered or clustered*
- Whether *missingness looks related to another variable* (which matters for how you handle it later)
:::
:::
:::
:::

# Missingness statistics {visibility="hidden"}

## Missing values in a dataset

- `n_miss()` from `naniar` counts the total number of missing values **across an entire dataset**

```{r}
n_miss(hrs_00)
```

- This is the total count of `NA` cells across *every* column combined
- `n_complete()` is the complement of `n_miss()`
    - It counts the total number of non-missing values across the dataset


 

- `prop_miss()` gives the proportion of all cells in the dataset that are missing

```{r}
prop_miss(hrs_00)
```

- Useful as a single, high-level number to report on the overall proportion missingness of your dataset

## Missing values in a single variable

- We can use `n_miss()` to count the number of missing values in a single variable

```{r}
n_miss(hrs_00$srh)
```

- This is the count of `NA` cells in self-reported health


 

- We can also use `prop_miss()` to get the proportion of missing values in a single variable

```{r}
prop_miss(hrs_00$srh)
```

- This is the proportion of `NA` cells in self-reported health




## Missingness by each variable, for every variable

- `miss_var_summary()` gives one row per variable, with the count and percent missing
- Piping the result into `gt()` makes it a nicer table to read

```{r}
hrs_00 |>
  miss_var_summary() |>
  gt() |>
  tab_options(table.font.size = px(40))
```

- The output is sorted from most to least missing, which makes it easy to spot the variables with the most missing data

# Visualizing missing data {visibility="hidden"}

## `vis_miss()`: visualizing missing data across the dataset

- This plot helps identify whether:
  - There are **variables** with a lot of missing values (in columns of plot)
  - **Observations** with a lot of missing values (in rows of plot)

::: columns
::: {.column width="40%"}

 

```{r}
#| eval: false
hrs_00 |>
  vis_miss() + #<1> 
  theme(axis.text.x = element_text(angle = 20, hjust = 0)) #<2>
```

1. `vis_miss()` is a function with a lot of built-in defaults that will produce a plot showing the missingness of the dataset
2. I made some changes to the `theme()` to make the variable names easier to read
:::
::: {.column width="60%"}
```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7

hrs_00 |>
  vis_miss() +
  theme_bw(base_size = 30) +
  theme(axis.text.x = element_text(angle = 20, hjust = 0))
```
:::
:::

## `gg_miss_fct()`: missingness heatmap by group

- This plot is useful for comparing missingness rates **stratified by a variable**
  - Potentially detect patterns in missingness
- Specify the variable to stratify by in the `fct =` option of `gg_miss_fct()`
  - Each **column** (x-axis) is a **different level of the stratifying variable**
  - Each **row** (y-axis) is a **different variable in the dataset**

::: columns
::: {.column width="40%"}

::: blue
::: blue-cont
**Example:** percent missingness stratified by `srh` (self-reported health)
:::
:::

```{r}
#| eval: false

hrs_00 |>
  gg_miss_fct(fct = srh) + #<1>
  theme_bw(base_size = 28) + #<2>
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) #<3>
```

1. `gg_miss_fct()` is a function that produces a heatmap of missingness stratified by a factor variable. We are stratifying by self-reported health (`srh`) in this example
2. I'm changing the background and font size to make the plot easier to read
3. I'm rotating the x-axis labels to make them easier to read
:::
::: {.column width="60%"}
```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 6

hrs_00 |>
  gg_miss_fct(fct = srh) +
  theme_bw(base_size = 28) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```
:::
:::

## `gg_miss_upset()`: missingness intersections (1/2)

- `gg_miss_upset()` shows the most common combinations (sets) of missing variables

::: columns
::: {.column width="30%"}
```{r}
#| eval: false

hrs_00 |>
  gg_miss_upset( #<1>
    nsets = 10, #<2>
    text.scale = 3 #<3>
    ) #<1> 
```

1. `gg_miss_upset()` is a function that produces an UpSet plot of missingness intersections
2. `nsets = 10` limits the plot to the 10 largest combinations of co-missing variables, which keeps the plot readable
3. `text.scale = 3` increases the size of the text in the plot, which makes it easier to read

:::
::: {.column width="70%"}
```{r}
#| echo: false
#| fig-width: 15
#| fig-height: 9

hrs_00 |>
  gg_miss_upset(
    nsets = 10, 
    text.scale = 3
    ) 
```
:::
:::



## `gg_miss_upset()`: missingness intersections (2/2)

- `gg_miss_upset()` shows the most common combinations (sets) of missing variables

::: columns
::: {.column width="30%"}
```{r}
#| eval: false

hrs_00 |>
  gg_miss_upset( #<1>
    nsets = 10, #<2>
    text.scale = 3 #<3>
    ) #<1> 
```

1. `gg_miss_upset()` is a function that produces an UpSet plot of missingness intersections
2. `nsets = 10` limits the plot to the 10 largest combinations of co-missing variables, which keeps the plot readable
3. `text.scale = 3` increases the size of the text in the plot, which makes it easier to read

- [**Main bar plot:**]{style="color: #5BAFF8;"} ordered by the number of observations with specific set of missing variables
- [**Bottom dots with lines:**]{style="color: #459B99;"} combination of variables missing
- [**Bottom, sideways barplot:**]{style="color: #C83532;"} number of observations missing for each variable, regardless of combination
:::
::: {.column width="70%"}
```{r}
#| echo: false
#| fig-width: 15
#| fig-height: 9

hrs_00 |>
  gg_miss_upset(
    nsets = 10, 
    text.scale = 3
    ) 
```
:::
:::


## `gg_miss_upset()`: What am I taking away from this plot?

::: columns
::: {.column width="60%"}
```{r}
#| echo: false
#| fig-width: 15
#| fig-height: 10

hrs_00 |>
  gg_miss_upset(
    nsets = 10, 
    text.scale = 3
    ) 
```
:::
::: {.column width="40%"}
- The most common combination of missing variables is `race_original` and `ed` with 465 observations
  - There are not many observations that have one of these missing without the other
  - There's 36 (34 + 2) observations that have `ed` missing but not `race_original`
  - There's 15 (12 + 3) observations that have `race_original` missing but not `ed`
  
- Height seems to be mostly missing on its own
  - 46 out of the 55 observations with height missing have no other variables missing
:::
:::

# Wrap-up {visibility="hidden"}

## Wrap-up

- We learned a few functions within `naniar` that help us summarize and visualize missingness in a dataset

 

- `n_miss()` and `prop_miss()` give quick, high-level counts of missingness across a whole dataset
- `miss_var_summary()` gives a tidy, sortable summary of missingness by variable
- `vis_miss()` gives a bird's-eye view of missingness across every row and column
- `gg_miss_fct()` compares missingness rates across groups
- `gg_miss_upset()` reveals which variables tend to be missing *together*


## Resources

- [Exploring missing values in `naniar`: interactive tutorial](https://allisonhorst.shinyapps.io/missingexplorer/)
- [`{naniar}` package website](http://naniar.njtierney.com/)
- [Missing values chapter in *R for Data Science (2e)*](https://r4ds.hadley.nz/missing-values.html)
- [Dealing with missing data in the *Epidemiologist R Handbook*](https://epirhandbook.com/en/missing-data.html)

