---
title: "Data transformation: Columns"
author: "Nicky Wakim"
title-slide-attributes:
    data-background-color: "#3070BF"
format: 
  revealjs:
    theme: "../simple_NW.scss"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    width: 1955
    height: 1100
    footer: "Data transformation: Columns"
    html-math-method: mathjax
    highlight-style: ayu
    code-annotations: hover
execute:
  echo: true
  freeze: auto
---

```{r}
#| label: "setup" 
#| include: false
#| message: false
#| warning: false

pacman::p_load(
  here, 
  tidyverse, 
  writexl, 
  rstatix, 
  janitor, 
  naniar,
  gt,
  gtsummary,
  readxl, 
  rio
)
```

```{r}
#| echo: false
hrs_data_rds <- import(here("data", "hrs_data.rds"))
hrs_00 <- clean_names(hrs_data_rds) |>
  relocate(race_original, sex, .after = degree) |>
  relocate(age_yr, ed, cesd, .after = birthdate)
```

## Data transformations on columns

- We can use columns to create new variables, subset features, and reorganize a dataset
- There are many functions that can be used to transform columns
- We will cover the most column functions:
  - `select()`
  - `relocate()`
  - `mutate()`
  - `rename()`

## Some notes before we get into it

::: pink
::: pink-ttl
Note 1: Assigning new objects
:::
::: pink-cont
My code in this lesson is a mix of assigning new objects and not assigning new objects. I do not assign new objects show you the output of the function without extra steps. 

- **For most of your work, you will want to assign your output to a new object.**
:::
:::

::: orange
::: orange-ttl
Note 2: Naming conventions
:::
::: orange-cont
- I use `hrs_00`, `hrs_01`, etc. to indicate the order of my transformations
- I use descriptive names for actions I take 
  - Example: `hrs_01_vars_subset` indicates that this is the 1st saved object I make and the action I take is subsetting variables (columns)
:::
:::

# `select()`

## Subsetting columns using `select()`

- The `select()` function allows us to isolate specific variables or columns from our dataset

 

- Why use `select()`?
  - Drops unneeded information to **maximize processing speed and visual clarity**
  - Isolates specific blocks of data for focused modeling or visualization
  
 

::: pink
::: pink-cont
Note: I will use `select()` in this lesson to make the output easier to read (with less columns)
:::
:::

## Example: selecting specific columns

- Let's keep only the following columns from `hrs_00`: `id`, `age_mo`, `income`, `diab`, `lung`, `drink`, `race_original`, and `sex`
  - Column names do NOT need quotes in `select()`

```{r}
hrs_01_vars_subset = hrs_00 |> 
  select(id, age_mo, income, diab, lung, drink, race_original, sex) # <1>
```

1.  We provide the exact column names we want to keep, separated by commas. All other columns will be dropped from the output.

::: columns
::: {.column width="48%"}
[Original dataset:]{style="color:#C83532"}

```{r}
tibble(hrs_00)
```
:::

::: {.column width="4%"}
:::

::: {.column width="48%"}
[Dataset with selected columns:]{style="color:#459B99"}
```{r}
tibble(hrs_01_vars_subset)
```
:::
:::

## Example: breaking code into multiple lines

- If code passes the vertical line in your editor pane, you can break it into multiple lines for readability

::: columns
::: column
```{r}
#| eval: false
hrs_01_vars_subset = hrs_00 |> 
  select(id, age_mo, income, 
         diab, lung, drink, 
         race_original, sex)
```
:::
::: column
```{r}
#| eval: false
hrs_01_vars_subset = hrs_00 |> 
  select(
    id, 
    age_mo, 
    income, 
    diab, 
    lung, 
    drink, 
    race_original, 
    sex
    )
```
:::
:::

## Example: removing specific columns

- If we want to keep everything **except** a few specified variables, we use a minus sign (`-`)

```{r}
hrs_01_vars_subset |> select(-drink, -lung) # <1>
```

1.  Placing `-` in front of `drink` and `lung` explicitly drops them, while retaining every other original variable.

## Example: selecting consecutive columns

- We can select a continuous block of columns using the `:` operator

```{r}
hrs_01_vars_subset |> select(age_mo:drink) # <1>
```

1.  This keeps `age_yrs`, `drink`, and every single column positioned between them in the data frame. **Make sure you know your column order before using this!**

## Distinguishing `filter()` vs `select()`

- Both `filter()` and `select()` are used to subset data, but they operate on different dimensions of the dataset

 

- `filter()` is used to subset [**rows**]{style="color: #5BAFF8;"} based on specific conditions
  - Think of this as filtering the dataset to keep only certain observations (e.g., only people who are 65+ years old)
  - We use this if we need a target population for analysis

 

- `select()` is used to subset [**columns**]{style="color: #E75B5C;"} based on specific names or positions
  - Think of this as selecting only the variables we need for analysis (e.g., only the variables we need to calculate a new variable)
  - We use this if we want to reduce the number of variables in our dataset for clarity or speed
  
# `relocate()`

## Reordering columns using `relocate()`

- `relocate()` moves columns to a different horizontal position in your dataset
- By default, variables sent into `relocate()` move directly to the **very front** of the dataset

```{r}
hrs_02_reloc <- hrs_01_vars_subset |> relocate(lung, sex) # <1>
```

1.  Moves `lung` and `sex` to the very front of the dataset, before all other columns.

 

::: columns
::: {.column width="48%"}
[Original subsetted dataset:]{style="color:#C83532"}

```{r}
hrs_01_vars_subset
```
:::

::: {.column width="4%"}
:::

::: {.column width="48%"}
[Dataset with `lung` and `sex` relocated to the front:]{style="color:#459B99"}
```{r}
hrs_02_reloc
```
:::
:::

## `relocate()` + specify location

- You can also specify where you want to move the column(s) using `.before` or `.after` arguments

```{r}
hrs_03_reloc <- hrs_01_vars_subset |> relocate(income, .after = id)       # <1>
hrs_04_reloc <- hrs_01_vars_subset |> relocate(income, .before = lung)        # <2>
```

1.  Moves `income` to place it directly after the `id` column.
2.  Moves `income` to place it right before the `lung`` column.

 

::: columns
::: {.column width="48%"}
[Original dataset:]{style="color:#C83532"}

```{r}
hrs_01_vars_subset
```
:::

::: {.column width="4%"}
:::

::: {.column width="48%"}
[Dataset with `income` relocated after `id`:]{style="color:#459B99"}
```{r}
hrs_03_reloc
```
:::
:::

# `mutate()`

![[Artwork by \@allison_horst](https://allisonhorst.com/everything-else){target="_blank"}](../img_slides/horst_mutate.jpg){fig-align="center"}


## `mutate()`: transform and create variables

- `mutate()` allows us to transform existing columns and/or add new columns to the dataset

 

- Think of this like adding a formula in Excel to calculate the value of a new column based on previous columns

 

- `mutate()` is useful for:
  - Converting the units of one column to new units (e.g., months to years)
  - Performing arithmetic using multiple columns
  - Changing variable types or recoding categories

## Example: calculating a new variable

- Let's convert `age_mo` into years by dividing it by 12
- We will name this new variable `age_yrs`

```{r}
hrs_04_reloc |> 
  mutate(age_yrs = age_mo / 12) |> # <1>
  relocate(age_yrs, .after = age_mo) # <2>
```

1.  **Notice we use a single `=` inside mutate**, not `==` or `<-`. This sets our new column name equal to the mathematical operation.
2.  We use `relocate()` to move the new column `age_yrs` to a more logical position in the dataset, right after `age_mo`. (And also so I can show you the change made.)

## Example: calculating multiple columns

- We can create multiple new columns within a single `mutate()` function by separating them with a comma

```{r}
hrs_05_mutate <- hrs_04_reloc |> # <1>
  mutate(
    age_yrs = age_mo / 12, # <2>
    age_decades = age_yrs / 10,  # <3>
    income_rounded = round(income, 0) # <4>
  )
```

1.  We use the assignment operator `<-` to save the output into a brand new data frame called `hrs_05_mutate`. I've already used up to `hrs_04_reloc` in the previous examples. 
2.  First, we create the `age_yrs` column.
3.  Second, we create `age_decades` using the `age_yrs` variable we *just* defined in the step right above it!
4. Third, we create `income_rounded` by rounding the `income` variable to the nearest whole dollar.


::: columns
::: {.column width="48%"}
[Original dataset:]{style="color:#C83532"}

```{r}
#| echo: false

hrs_01_vars_subset
```
:::

::: {.column width="4%"}
:::

::: {.column width="48%"}
```{r}
#| code-fold: true
#| code-summary: "Dataset with new variables:"

hrs_05_mutate |> 
  relocate(age_yrs, age_decades, .after = age_mo) |>
  relocate(income_rounded, .after = income)
```
:::
:::

# `rename()`

## Renaming with `rename()`

-   You can rename columns with `select()` but it's a bit easier to do this with the `rename()` function.

-   You just need to remember `new_name = old_name` ordering.

-   For example, I want to highlight to myself the limited representation in the `race_original` and `sex` variables
    - I'll rename them to `race_3` and `sex_2` to indicate that they are limited to 3 and 2 categories, respectively

```{r}
hrs_06_rename <- hrs_05_mutate |> 
  rename(
    race_3 = race_original, #<1> 
    sex_2 = sex
    )
names(hrs_06_rename)
```

1.  The new name is on the left, and the old name is on the right. 

::: red
::: red-cont
If our variable names were something like "Age (years)", we would need to write `age_yrs = 'Age (years)'` with single quotes around the old name because of the spaces and parentheses.
:::
:::

## Wrap-up

- We learned three functions that allow us to transform columns in our dataset:
  - `relocate()`
    - Reorder columns to a different spot in the dataset
  - `select()`
    - Subset specific columns to keep or drop
  - `mutate()`
    - Transform existing columns and/or create new columns
  - `rename()`
    - Rename columns with new names
